---
title: "bulkanalysis_demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bulkanalysis_demo_ver2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 12,
  fig.height = 10,
  out.width = "100%"
)
```

```{r setup}
library(bulkanalysis)
```

# Introduction

The purpose of the bulkanalysis package is to streamline the downstream analysis of bulk RNA-seq data by providing a collection of commonly used analysis and visualizations in a single package. The package was developed as a wrapper around statistical and visualization methods that I have frequently encountered during my experience analyzing bulk RNA-seq data. Rather than introducing new statistical methods, bulkanalysis is designed to reduce repetitive coding and provide a consistent workflow for obtaining commonly used downstream results. This package performs techniques such as differential expression via [edgeR](https://web.stanford.edu/class/bios221/labs/rnaseq/lab_4_rnaseq.html) with traditional outputs such as heatmaps, volcano plots. As well as over-representation analysis via [clusterProfiler](https://yulab-smu.top/biomedical-knowledge-mining-book/enrichplot.html) using [Fishcher's Exact Test](https://www.youtube.com/watch?v=udyAvvaMjfM) with summarized output visuals (As of 9/17/26).

The raw results and output visuals are accessible by either executing the individual functions or through the `bulkanalysis::run_de_pipeline()}`'s nested list output.This vignette below will go into detail on the inputs necessary for the `bulkanalysis::run_de_pipeline()` and the outputs of functions. The overall goal of this package is to save time on initial biological inferences of bulk-RNAseq experiments by cutting out repetitive coding, visualization building, and parsing through large noise of large statistical results. Delivering summarized results for faster interpretations of your data. By combining these steps into reusable functions, the package provides summarized results that can be used to more efficiently explore potential biological patterns in the data. These results are intended to support initial biological interpretation rather than replace more comprehensive statistical or biological investigation.

Input Data First we need to format a random bulk-RNAseq dataset that has been pre-processed (from FASTQ files), but unnormalized counts to enter into a [DESeqDataSet](https://www.rdocumentation.org/packages/RNAseqQC/versions/0.2.1/topics/make_dds) via `RNAseqQC::make_dds()`. This is because make_dds() does not accept continuous counts. Which requires at minimum a count matrix in the `matrix` class from base-r and meta-data table with following columns:

`sample_names`: The names of the samples of your bulk-RNAseq experiment in the same exact arrangement and character format (capitlaize/lower_case).

`sample_condition`: which tells the experimental condition of your samples.

These inputs can then be provided to `bulkanalysis::edgeR_diffexp()` to perform pairwise differential expression analysis between experimental conditions. A reference condition is specified so that the resulting comparisons are interpreted relative to a defined baseline. In many experiments, this reference may be a condition such as control or healthy, although the appropriate reference depends on the experimental design.

# Bulk-RNASeq Formatting

This section will show you how to prepare a random dataset from [GEO Query](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE202805) for the bulkanalysis package functions. GEO202805 will be used to prepare a DESeqDataSet object compatible with the rest of functions in the bulk analysis package.

**GEO202805 Summary**

RNA sequencing was performed on a total of 80 PBMC samples derived from healthy controls (n = 10), acute-Mild (n = 4), Acute-Moderate (n = 6), Acute-Severe (n = 32) and Convalescent = 28. For this

**Citation**: Minghan,.L , Yuqing,.S, Yizhou,.T, Yuefan,.L, Weidong,.T, " Global Transcriptomic analysis of PBMC derived from COVID-19 acute-severe patients and convalescents", (May, 2022),PRJNA837285; GEO: GSE202805.

### Downloading GEOquery Data

```{r, message=FALSE, include = FALSE}
library(GEOquery)
#read in the whole count matrices and meta data

gse_202805 <- getGEO("GSE202805", GSEMatrix = TRUE)

#divide the ExpressionSet into the expression matrix and meta data

gse_202805 <- gse_202805[[1]]

#expression matrix

expr_202805 <- exprs(gse_202805)

#meta-data
meta_202805 <- pData(gse_202805)

#showing first five rows of the meta-data and the expression samples
head(expr_202805, n = 5)
head(meta_202805, n = 5)
#download the data into the data directory
getGEOSuppFiles("GSE202805")
```

Check meta data for which groups you would like set the comparisons to.For this demonstration we will be binning the all the acute covid-19 samples (mild, moderate, severe). In your metadata, you can partition each of these labels in a more granular fashion.

```{r}
#checking if path exists
path <- "./GSE202805/GSE202805_genecounts_20220406.txt.gz"

counts <- utils::read.delim(gzfile(path))

colnames(counts) <- base::sub("\\.bam$", "", colnames(counts))
#Some dataset will still come with the sorted bam file that outputs from command line functions such as featureCounts (subread)
#The columns being deleted are not necessary unless we are integrating ChIP or ATAC data.
counts <- counts %>% dplyr::select(-c(Chr, Start, End, Strand, Length))
# contains the count vectors of the 8 samples
#Some dataset will still come with the sorted bam file that outputs from command line functions such as featureCounts (subread)
counts[1:10, 1:10]
```

Some datasets will still come within the sorted bam file that outputs from command line functions such as featureCounts (subread). In case that happens use the select(-c()) to get rid of columns such as Chr, Start, End, Strand, Length. Which useful if you have genomic (ChIP)or epigenomic (ATAC). However we will have to use tidyverse on the metadata table to only the basic relevant columns.

# Meta-Data Editing

```{r}
library(data.table)
library(tidyverse)
library(stringr)


meta_data <- meta_202805

meta_data <- meta_data %>%
  rownames_to_column(var = "samples") %>%
  dplyr::select(samples, title, characteristics_ch1) %>%
  mutate(characteristics_ch1 = gsub("group:", "", characteristics_ch1)) %>%
  dplyr::select(-c(samples))

mapping <- c(meta_data$characteristics_ch1)
names(mapping) <- meta_data$title

colnames(counts) <- mapping[colnames(counts)] 

colnames(counts) <- paste(names(mapping), colnames(counts), sep = "_")

colnames(counts)[1] <- "geneid"


# Character Specific Clean Up
counts$geneid <- gsub("\\..*", "", counts$geneid)
colnames(counts) <- gsub(" ", "", colnames(counts))

# So this is a crucial part to remember lol is that acute-covid19, convalescant covid
#added waay late into the game after first round of harmonization and integration
colnames(counts) <- gsub("acute", "covid19", colnames(counts))

# MetaData construction
metadata <- data.frame(
  sample = colnames(counts),
  condition = dplyr::case_when(
    grepl("_covid19$", colnames(counts)) ~ "covid19",
    grepl("_convalescent$", colnames(counts)) ~ "convalescent",
    grepl("healthy$", colnames(counts)) ~ "healthy",
    TRUE ~ NA_character_
  ) 
) %>% dplyr::slice(., 2:nrow(.)) %>%
  dplyr::mutate(condition = as.factor(condition)) %>%
  dplyr::mutate(sample = as.factor(sample))

```

```{r}
# summarizing replicated genes across all samples
counts <- counts %>%
  group_by(geneid) %>%
  summarise(across(where(is.numeric), sum))

counts <- counts %>% column_to_rownames(var = "geneid")


metadata[] <- lapply(metadata, factor)

```

```{r}
counts[]
metadata[10, 10]
```

# Making DDS Object

make_dds(): Make DESeqDataSet from counts matrix and metadata.

Per transcript counts of RNA-seq exhibit a mean-variance relationship ([hetroscedastic](https://www.biostars.org/p/9587730/)). However many tests within RNA-seq pipeline require non-hetroscedastic data. Normalizations are require to smooth the variation of the data. [Normalization](https://hbctraining.github.io/DGE_workshop/lessons/02_DGE_count_normalization.html) scales the raw count of values to account for the technical batch effects between samples. This way expression level are comparable between and within samples.

**`vst()` (wrapper for `DESeq2::varianceStabilizingTransformation()`)**: variance stabilizing transformation transforms normalizes count data to reduce the dependence between the mean and variance. This produces an approximately homoscedastic representation of the data, making it more appropriate for exploratory analyses based on distances or correlations, such as PCA and clustering. The transformed values are intended for exploratory analysis and visualization rather than differential-expression testing.The result is applied to the `bulkanalysis::baseline_dimreduction()` function.

**Log normalization**: The typical distribution of bulk-RNAseq data follows a Poisson (Var = mean) or a Negative Binomial Distribution (Var \> Mean) For data containing a large amount of zeros, the log transformation serves to minimize the frequency of zeroes relative to transcripts with counts \>= 1. This will help to visualize the data though a normal distribution. This is typcial are typical for scRNA-Seq and ST data. To avoid taking the log of 0, it is common to add a +1 pseudo value.

More information of [Normalization](https://www.youtube.com/watch?v=u3395drEfrs&list=PLg0JKLUfmkdkIr-9hvFLY1O1JRqOdjTSW)

**Note**: These transformations are not used within the differential expression analysis. This is because packages such as EdgeR and DESeq2 are built to take into account for the typical Poisson or Negative Binomial distribution.

```{r}

library(DESeq2)
library(RNAseqQC)
library(AnnotationHub)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)


# Unflitered Counts Saved to a DeSeqDataSet (dds object)
covid19_ds1 <- make_dds(counts = counts , metadata = metadata, ah_record = "AH116291")
#renaming counts to raw counts
assayNames(covid19_ds1)[1] <- "counts"

#variance stabilized Raw counts
assays(covid19_ds1)[[2]] <- vst(covid19_ds1@assays@data$counts)
assayNames(covid19_ds1)[2] <- "var_stable"
#log_transformed (log normalized counts)
assays(covid19_ds1)[[3]] <- log10(covid19_ds1@assays@data$counts + 1)
assayNames(covid19_ds1)[3] <- "log_counts"
```

```{r}
# saveRDS does not work in the vignette build
#saveRDS(covid19_ds1,"./vignette_data/gse202805_test_data.rds")
```

# Bulk Analysis Demonstration

standardization of meta-data labels.

```{r}
# In some scenarios, you will need to standardize your condition
covid19_ds1@colData$condition <- covid19_ds1@colData$condition %>% tolower()
```

### EdgeR Based Differential Expression

Performs EdgeR based differential expression with limma support of each covariate specified in your meta-data with user set reference variable for pairwise differential expression of each sample condition within the submitted meta-data of your formatted Deseqdataset object. This is a standard pipeline sourced from [Daniel Neves and Daniel Sobral, "Using DESeq2 and edgeR in R", (April, 2018) vignette](https://gtpb.github.io/ADER18F/pages/tutorial1.html)

1\. Extracts Raw counts

2.  filters for low count genes within your count matrix and keeps sufficiently large counts to be retained in a statistical analysis (`edgeR::filterByExpr(y)`). Filtering is performed with respect to the experimental design/replication structure, rather than using an arbitrary fixed count cutoff.

3.   `edgeR::normLibSizes(y)` calculates normalization factors, commonly using **TMM.**([TMM normalization](https://davetang.github.io/muse/edger.html) via `edgeR::normLibSizes(y)`). These factors adjust the effective library sizes so that samples are comparable despite differences in sequencing depth and RNA composition.

4.  Re-affirms reference level relative to the experimental covariate

5.  set design matrices as mean reference model expression = B1(reference)m + B2(experiemnt) for each pairwise combination of conditions in the metadata.

    `design <- model.matrix(~0+condition, data = y$samples).`the model estimates the mean expression associated with each condition, and contrasts between those coefficients define the desired comparisons. THIS IS NOT THE PRIMARY COMPARISON ONLY FOR DESIGN MATRIX CONTRSUCTION.

6.5: Estimates variability of each gne in the count data

6.  Set limma::makeContrats() for each comparison of contrasts vectors to be applied in the [`edgerR::glmQLFit()`](https://www.rdocumentation.org/packages/edgeR/versions/3.14.0/topics/glmQLFit) function.

For each contrast (comparison (reference vs x) and vector)

7\. edgerR::glmQLFit() which fits the negative binomial generalised linear model using quaisi-liklihood methods and determines whether the observes group difference are larger than expected variability.

8.Applies either Family Wise Error('bonferonni') or False Discovery Rate('BH') pvalue correction

9.  Applies user filter of initial p.adjust value. Default is no filter applied.

**Output**: A list object of data frames containing each possible combination of binary \#' comparisons (control vs. experimental group) within the supplied metadat

```{r}
dds_object <- covid19_ds1
#Add the condition column name
condition_col = "condition"
#set a refreence group name
ref_group_name = "healthy"


#Performs inital differential expression on your DDS Object's raw counts dataframe
# The use give the condition column from the metadata to make the test parameters
# and the control group
results_dds1 <- edgeR_diffexp(covid19_ds1, 
                              condition_col,  
                              ref_group_name,
                              init_pval_cutoff = 0.5,
                              adjust.method = "bonferroni")
```

Result is the list of valid binary comparisons within the data using EdgeR.

See the `edgeR_diffexp()` documentation for more details.

```{r}
healthy_covid19_results <- results_dds1$`healthy-covid19`
```

```{r}
healthy_covid19_results <- results_dds1$`healthy-covid19`

results_dds1$`healthy-convalescent`[1:10, 1:ncol(healthy_covid19_results)]

```

# **Running the whole DE Downstream pipeline**

Inputs: the DDS object, Results of the EdgeR Differential Expression analysis

Arguments:

- `dds_object`:A formatted DESeq2 object where each sample and its counts are aligned with the respective metadata in `@colData`.

- `edgeR_results`: A dataframe of differential expression results (minimum columns: log2fc, pvalue, p.adjust) **DEG statisitcal inputs**

- `deg_log2fc_thresh`: absolute value cutoff of log2fc

- `deg_pval_adj_thresh`: pvalue cutoff for initial log2fc values

**Volcano Plots Arguments**

- `vol_plot_lower_xlim`: Numeric value specifying the lower limit of the x-axis. Should be less than 0. For example, `-5`.

- `vol_plot_upper_xlim`: Numeric value specifying the upper limit of the x-axis. Should be greater than 0. For example, `5`.

- `cartesian_step`: Numeric value specifying the interval between x-axis ticks.

**ORA Arguments**

- `ora_pval_adj_threshold`: pvalue cutoff of the over-representation test

- `ora_convert_ids`: Boolean option to convert the gene IDs from Ensemble to External format

- `orgdb`: An organism-specific annotation database used to facilitate gene identifier conversion, such as org.Hs.eg.db for human or org.Mm.eg.db for Mus Musculous.

- `ensembl_dataset`: Character string specifying the Ensembl dataset used for annotation. For example, `"hsapiens_gene_ensembl"` for human genes.

```{r}
library(org.Hs.eg.db)
covid19_ds1_analysis <- run_de_pipeline(
                                covid19_ds1, 
                                results_dds1,
                                deg_log2fc_thresh = 1, 
                                deg_pval_adj_thresh = 0.05, 
                                vol_plot_lower_xlim = -10, 
                                vol_plot_upper_xlim = 10, 
                                cartesian_step = 5, 
                                ora_pval_adj_threshold = 0.10,
                                orgdb = org.Hs.eg.db,
                                ensembl_dataset = "hsapiens_gene_ensembl"
                              )
```

## Outputs

The outputs of the run_de_pipeline are described below

`filtered_results`: The initial filtered result of DEGs log2fc and pval_adj_threshold.

**Columns:**

\`log2FC\`: The log2-scaled fold change of a given gene across the \#' specified samples.

`P-value`: The probability of obtaining data as extreme as the \#' observed values assuming that the null hypothesis is true.

`Q-value (FDR)`: The proportion of false positives after accounting \#' for multiple testing.

```{r}
library(glue)
demo_results <-covid19_ds1_analysis$filtered_results$`healthy-covid19`

glue("Total DEGs in healthy vs covid-19 comparison: {dim(demo_results)[1]}")

demo_results[1:10, 1:ncol(demo_results)]
```

run_de_pipeline outputs

`volcano_plots`: visualization of the filtered results on a volcano plot

`heatmap`:

`ora_up_raw`: Result of the over-representation analysis on upregulated DEGs via clusterProfiler `unpacked_results_up`: unpacked results table consisting of a nested list of ORA results on the upregulated DEGs

`ora_down_raw`: Result of the over-represenation analysis on downregulated DEGs via clusterProfiler `unpacked_results_down`: unpacked results table consisting of a nested list of ORA results on the down-regulated DEGs

## Volcano Plot

Uses ComplexHeatMap to compare z-scaled expression of DEGs within the filtered results

```{r, fig.height = 7, fig.width=10, dpi = 300}
covid19_ds1_analysis$volcano_plots$`healthy-covid19`
```

### Heatmaps

Uses ComplexHeatMap to compare z-scaled expression of DEGs from the filtered results table.

```{r, fig.height = 10, fig.width = 7, dpi=300}
covid19_ds1_analysis$heatmap$`healthy-covid19`
```

### ORA DotPlot Summary for Up Regulated and DownRegulated DEGs

Summarizes the the statistical significance and the count of each DEG in an over-represented term.

```{r, fig.height = 5, fig.width = 10, dpi=300}
covid19_ds1_analysis$ora_down_raw$enrichgo_visuals$`healthy-covid19`
```

# Future Work

Implement GSEA

Protein-Protein Interaction Network

New ORA visuals: Radial Upset Plot, ORA network visual for centrality.

# Citations
