---
title: "bulkanalysis_demo_ver2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bulkanalysis_demo_ver2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bulkanalysis)
```

# Bulk-RNASeq Formatting

The purpose is to show how to prepare a random dataset from [GEO Query](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE133432) for the bulk-analysis package. GEO202805 will be used to prepare a Deseq2dataset object compatable with the rest of functions in the bulk analysis package.

**GEO202805 Summary**

RNA sequencing was performed on a total of 80 PBMC samples derived from healthy controls (n = 10), acute-Mild (n = 4), Acute-Moderate (n = 6), Acute-Severe (n = 32) and Convalescent = 28

```{r}
library(GEOquery)
#read in the whole count matrices and meta data

gse_202805 <- getGEO("GSE202805", GSEMatrix = TRUE)

#divide the ExpressionSet into the expression matrix and meta data

gse_202805 <- gse_202805[[1]]

#expression matrix

expr_202805 <- exprs(gse_202805)

#meta-data
meta_202805 <- pData(gse_202805)


head(expr_202805)
head(meta_202805)
#download the data into the data directory
getGEOSuppFiles("GSE202805")
```

Check meta data for which groups you would like set the comparisons to.

```{r}
#checking if path exists
path <- "./GSE202805/GSE202805_genecounts_20220406.txt.gz"

counts <- utils::read.delim(gzfile(path))

colnames(counts) <- base::sub("\\.bam$", "", colnames(counts))
#Some dataset will still come with the sorted bam file that outputs from command line functions such as featureCounts (subread)
counts <- counts %>% dplyr::select(-c(Chr, Start, End, Strand, Length))
# contains the count vecotrs of the 81 samples
counts 
#Some dataset will still come with the sorted bam file that outputs from command line functions such as featureCounts (subread)

```

Some datasets will still come within the sorted bam file that outputs from command line functions such as featureCounts (subread). In case that happens use the select(-c()) to get rid of columns such as Chr, Start, End, Strand, Length. Which useful if you have epigenomic data such as ChIP or ATAC data. However we will have to use tidyverse on the metadata table to only the relevant columns

Meta-Data Editing

```{r}
library(data.table)
library(tidyverse)
library(stringr)


meta_data <- meta_202805
meta_data
meta_data <- meta_data %>%
  rownames_to_column(var = "samples") %>%
  dplyr::select(samples, title, characteristics_ch1) %>%
  mutate(characteristics_ch1 = gsub("group:", "", characteristics_ch1)) %>%
  dplyr::select(-c(samples))

mapping <- c(meta_data$characteristics_ch1)
names(mapping) <- meta_data$title

colnames(counts) <- mapping[colnames(counts)] 

colnames(counts) <- paste(names(mapping), colnames(counts), sep = "_")

colnames(counts)[1] <- "geneid"


# Character Specific Clean Up
counts$geneid <- gsub("\\..*", "", counts$geneid)
colnames(counts) <- gsub(" ", "", colnames(counts))

# So this is a crucial part to remember lol is that acute-covid19, convalescant covid
#added waay late into the game after first round of harmonization and integration
colnames(counts) <- gsub("acute", "covid19", colnames(counts))

# MetaData construction
metadata <- data.frame(
  sample = colnames(counts),
  condition = dplyr::case_when(
    grepl("_covid19$", colnames(counts)) ~ "covid19",
    grepl("_convalescent$", colnames(counts)) ~ "convalescent",
    grepl("healthy$", colnames(counts)) ~ "healthy",
    TRUE ~ NA_character_
  ) 
) %>% dplyr::slice(., 2:nrow(.)) %>%
  dplyr::mutate(condition = as.factor(condition)) %>%
  dplyr::mutate(sample = as.factor(sample))
```

```{r}
# summarizing replicated genes across all samples
counts <- counts %>%
  group_by(geneid) %>%
  summarise(across(where(is.numeric), sum))

counts <- counts %>% column_to_rownames(var = "geneid")


metadata[] <- lapply(metadata, factor)

```

```{r}
counts
metadata
```

### Making DDS Object

```{r}

library(DESeq2)
library(RNAseqQC)
library(AnnotationHub)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)


# Unflitered Counts Saved to a DeSeqDataSet (dds object)
covid19_ds1 <- make_dds(counts = counts , metadata = metadata, ah_record = "AH116291")
#renaming counts to raw counts
assayNames(covid19_ds1)[1] <- "counts"

#variance stabilized Raw counts
assays(covid19_ds1)[[2]] <- vst(covid19_ds1@assays@data$counts)
assayNames(covid19_ds1)[2] <- "var_stable"
#log_transformed (log normalized counts)
assays(covid19_ds1)[[3]] <- log10(covid19_ds1@assays@data$counts + 1)
assayNames(covid19_ds1)[3] <- "log_counts"
```

```{r}
# saveRDS does not work in the vignette build
#saveRDS(covid19_ds1,"./vignette_data/gse202805_test_data.rds")
```

Necessary QC is required please check other resources for bulk RNA-Seq QC tutorial and rationale

### Bulk Analysis Demonstration

```{r}
# In some scenarios, you will need to standardize your condition
covid19_ds1@colData$condition <- covid19_ds1@colData$condition %>% tolower()
```

### Edge R Based Differential Expression

```{r}
dds_object <- covid19_ds1
#Add the condition column name
condition_col = "condition"
#set a refreence group name
ref_group_name = "healthy"


#Performs inital differential expression on your DDS Object's raw counts dataframe
# The use give the condition column from the metadata to make the test parameters
# and the control group
results_dds1 <- edgeR_diffexp(covid19_ds1, 
                              condition_col,  
                              ref_group_name,
                              init_pval_cutoff = 0.5,
                              adjust.method = "bonferroni")
```

Result is the list of valid binary comparisons within the data using EdgeR.

See the `edgeR_diffexp()` documentation for more details \code{\link{edgeR_diffexp}}

```{r}
results_dds1$`healthy-convalescent`
```

**Running the whole DE Downstream pipeline**

Inputs: the DDS object, Results of the EdgeR Differential Expression analysis

Outputs:

```{r}
library(org.Hs.eg.db)
covid19_ds1_analysis <- run_de_pipeline(
                                covid19_ds1, 
                                results_dds1,
                                deg_log2fc_thresh = 1, 
                                deg_pval_adj_thresh = 0.05, 
                                vol_plot_lower_xlim = -10, 
                                vol_plot_upper_xlim = 10, 
                                cartesian_step = 5, 
                                ora_pval_adj_threshold = 0.10,
                                orgdb = org.Hs.eg.db,
                                ensembl_dataset = "hsapiens_gene_ensembl"
                              )
```

## Volcano Plot

```{r}
covid19_ds1_analysis$volcano_plots$`healthy-covid19`
```

### Heatmaps

```{r}
covid19_ds1_analysis$heatmap$`healthy-covid19`
```

### ORA DotPlot Summary for Up Regulated and DownRegulated DEGs

```{r}
covid19_ds1_analysis$ora_up_raw$enrichgo_visuals$`healthy-covid19`
covid19_ds1_analysis$ora_down_raw$enrichgo_visuals$`healthy-covid19`
```
