---
title: "Intra_Differential_Expression"
output:
  pdf_document: default
  html_document: default
---

# Read-In Libraries

```{r}
start.time <- Sys.time()
```

```{r, include =FALSE}
library(tidyverse)
library(glue)
library(devtools)
library(vroom)
library(GEOquery)
library(biomaRt)
library(dplyr)
library(org.Hs.eg.db)
library(DESeq2)
library(clusterProfiler)
library(RNAseqQC)
library(edgeR)
source("../Project_1_function_lab.R")
```

```{r}
covid19_ds1 <- readRDS("../post_processed_data/gse202805_covid19_ds1.rds")
covid19_ds2 <- readRDS("../post_processed_data/gse152641_covid19_ds2.rds")
covid19_ds3 <- readRDS("../post_processed_data/gse161731_covid19_ds3.rds")
covid19_ds4 <- readRDS("../post_processed_data/gse227116_covid19_ds4.rds")
covid19_ds5 <- readRDS("../post_processed_data/gse152418_covid19_ds5.rds")

covid19_dsint <- readRDS("../post_processed_data/covid_dds_12345.rds")


# Adding DataSet Moniceurs
colData(covid19_ds1)$ds_origin <- "GSE202805"
colData(covid19_ds2)$ds_origin <- "GSE152641"
colData(covid19_ds3)$ds_origin <- "GSE161731"
colData(covid19_ds4)$ds_origin <- "GSE227116"
colData(covid19_ds5)$ds_origin <- "GSE152418"
```

```{r}
covid19_ds1@colData$condition <- covid19_ds1@colData$condition %>% tolower()
covid19_ds2@colData$condition <- covid19_ds2@colData$condition %>% tolower()
covid19_ds3@colData$condition <- covid19_ds3@colData$condition %>% tolower() %>% gsub("-", "",.) %>% as.factor()

covid19_ds4@colData$condition <- covid19_ds4@colData$condition %>% tolower()

covid19_ds5@colData$condition <- covid19_ds5@colData$condition %>% tolower() %>% gsub("-", "",.)
```

```{r}
levels(covid19_ds3@colData$condition)
```

<https://olvtools.com/en/documents/edger>

# Part2_Project 1 Intra Diffferential Expression of the DataSets

### Perform EdgeR Differential Expression and plot a volcano plot for each condition comparison within the DDS object

```{r}

#Arguments
dds_object <- covid19_ds1
condition_col = "condition"
ref_group_name = "healthy"


#Performs inital differential expression on your DDS Object's raw counts dataframe
# The use give the condition column from the metadata to make the test parameters
# and the control group
results_dds1 <- edgeR_diffexp(covid19_ds1, 
                              condition_col,  
                              ref_group_name,
                              init_pval_cutoff = 0.5,
                              adjust.method = "bonferroni")



results_dds2 <- edgeR_diffexp(covid19_ds2, 
                              condition_col, 
                              ref_group_name,
                              init_pval_cutoff = 0.5,
                              adjust.method = "bonferroni")

results_dds3 <- edgeR_diffexp(covid19_ds3, 
                              condition_col, 
                              ref_group_name,
                              init_pval_cutoff = 0.5,
                              adjust.method = "bonferroni")

results_dds4 <- edgeR_diffexp(covid19_ds4, 
                              condition_col, 
                              ref_group_name,
                              init_pval_cutoff = 0.5,
                              adjust.method = "bonferroni")

results_dds5 <- edgeR_diffexp(covid19_ds5, 
                              condition_col, 
                              ref_group_name,
                              init_pval_cutoff = 0.5,
                              adjust.method = "bonferroni")
```

# **Covid-19 DS1 GSE202805 Downstream Analaysis**

```{r, include=FALSE}

covid19_ds1_analysis <- run_de_pipeline(
                                        covid19_ds1, 
                                        results_dds1,
                                        deg_log2fc_thresh = 1, 
                                        deg_pval_adj_thresh = 0.05, 
                                        vol_plot_lower_xlim = -10, 
                                        vol_plot_upper_xlim = 10, 
                                        cartesian_step = 5, 
                                        ora_pval_adj_threshold = 0.05
                                        )
#saveRDS(covid19_ds1_analysis, "./downstream_results/covid_ds1_gse202805")
```

### **Covid-19 DS1 GSE202805 significant DEGs abs(log2fc) \> 1 and pval_adj \< 0.05**

**healthy-convalalescent: \# DEGs total = 114**

Upregulated: 19

Down Regulated: 95

**Healthy-covid19: #DEGs total = 295**

UpRegulated: 61

Downregulated: 234

Convalescent-covid19: \# DEGs total = 1408

Upregulated: 535 Sig up

Downregulated: 87

### Displaying All the Results

```{r}
covid19_ds1_analysis$volcano_plots
covid19_ds1_analysis$heatmap
covid19_ds1_analysis$ora_down_raw$enrichgo_visuals
covid19_ds1_analysis$ora_up_raw$enrichgo_visuals
```

# **Covid-19 DS2 GSE**152641 **Downstream Analysis**

```{r}
covid19_ds2_analysis <- run_de_pipeline(covid19_ds2, 
                                        results_dds2,
                                        deg_log2fc_thresh = 1, 
                                        deg_pval_adj_thresh = 0.05, 
                                        vol_plot_lower_xlim = -10, 
                                        vol_plot_upper_xlim = 10, 
                                        cartesian_step = 5, 
                                        ora_pval_adj_threshold = 0.05)
```

```{r}
#saveRDS(covid19_ds1_analysis, "./downstream_results/covid_ds2_gse152641")
```

### Covid-19 DS2 GSE152641 **significant DEGs abs(log2fc) \> 1 and pval_adj \< 0.05**

**Healthy-covid19: #DEGs total = 591**

**UpRegulated: 112 sig up**

**Downregulated: 479 sig down**

### Displaying All the Results

```{r}
covid19_ds2_analysis$volcano_plots
covid19_ds2_analysis$heatmap
covid19_ds2_analysis$ora_down_raw$enrichgo_visuals
covid19_ds2_analysis$ora_up_raw$enrichgo_visuals
```

# Covid19- DS3 GSE161731 Downstream Analysis

```{r}
covid19_ds3_analysis <- run_de_pipeline(covid19_ds3, 
                                        results_dds3,
                                        deg_log2fc_thresh = 1, 
                                        deg_pval_adj_thresh = 0.05, 
                                        vol_plot_lower_xlim = -10, 
                                        vol_plot_upper_xlim = 10, 
                                        cartesian_step = 5, 
                                        ora_pval_adj_threshold = 0.5)
```

# **Covid-19 DS4 GSE**227116 **Downstream Analaysis**

```{r, include=FALSE}
covid19_ds4_analysis <- run_de_pipeline(covid19_ds4, 
                                        results_dds4,
                                        deg_log2fc_thresh = 1, 
                                        deg_pval_adj_thresh = 0.05, 
                                        vol_plot_lower_xlim = -10, 
                                        vol_plot_upper_xlim = 10, 
                                        cartesian_step = 5, 
                                        ora_pval_adj_threshold = 0.5)
```

```{r}
#saveRDS(covid19_ds2_analysis, "./downstream_results/covid_ds4_gse227116")
```

```{r}

```

### Displaying All the Results

```{r}
covid19_ds4_analysis$volcano_plots
covid19_ds4_analysis$heatmap
covid19_ds4_analysis$ora_down_raw$enrichgo_visuals
covid19_ds4_analysis$ora_up_raw$enrichgo_visuals

```

### Significant DEGs

**Covid-19 DS4 GSE227226 significant DEGs abs(log2fc) \> 1 and pval_adj \< 0.05**

**healthy vs convascent: total DEGs: 111**

**24 sig up, 87 sig down**

Over_Representated Pathways

Observation: There are Barely any over-represented pathways after inputting the 111 DEGs into enrichGO. This is consistent for both the pval threshold at 0.05 for both up and down regulated DEGs as input.

# Covid-19 DS5 GSE152418 Downstream Analysis

```{r}
covid19_ds5_analysis <- run_de_pipeline(covid19_ds5, 
                                        results_dds5,
                                        deg_log2fc_thresh = 1, 
                                        deg_pval_adj_thresh = 0.05, 
                                        vol_plot_lower_xlim = -10, 
                                        vol_plot_upper_xlim = 10, 
                                        cartesian_step = 5, 
                                        ora_pval_adj_threshold = 0.5)
```

### Dispalying Results

```{r}
covid19_ds5_analysis$volcano_plots
covid19_ds5_analysis$heatmap

```

# Integrated Covid Dataset (**GSE202805 + GSE**227116 + **GSE**152641) Downstreamt Analysis

```{r}
covid19_dsint <- readRDS("./post_processed_data/covid_dds_12345.rds")
```

```{r}
disease_states <- c(levels(covid19_dsint$condition))

comb = function(n, x) {
  factorial(n) / factorial(n-x) / factorial(x)
}

comb(length(disease_states), 2)
```

```{r, include =FASLE}
unique(covid19_dsint@colData$condition)
dds_object <- covid19_dsint 
condition_col = "condition"
ref_group_name = "healthy"
results_ddsint <- edgeR_diffexp(covid19_dsint,
                                condition_col,
                                ref_group_name)

covid19_ds124_analysis <- run_de_pipeline(covid19_dsint, 
                                        results_ddsint,
                                        deg_log2fc_thresh = 1, 
                                        deg_pval_adj_thresh = 0.05, 
                                        vol_plot_lower_xlim = -20, 
                                        vol_plot_upper_xlim = 20, 
                                        cartesian_step = 5, 
                                        ora_pval_adj_threshold = 0.05)
```

```{r}
#saveRDS(covid19_ds124_analysis, "./downstream_results/covid_ds124_integrated")
```

```{r}
if(!dir.exists("downstream_results")) dir.create("downstream_results", recursive = TRUE)

save(
  covid19_ds1_analysis,
  file = "./downstream_results/covid19_ds1_analysis.RData",
  compress = "gzip"
)

save(
  covid19_ds2_analysis,
  file = "./downstream_results/covid19_ds2_analysis.RData",
  compress = "gzip"
)

save(
  covid19_ds3_analysis,
  file = "./downstream_results/covid19_ds3_analysis.RData",
  compress = "gzip"
)

save(
  covid19_ds4_analysis,
  file = "./downstream_results/covid19_ds4_analysis.RData",
  compress = "gzip"
)

save(
  covid19_ds5_analysis,
  file = "./downstream_results/covid19_ds5_analysis.RData",
  compress = "gzip"
)

save(
  covid19_ds124_analysis,
  file = "./downstream_results/covid_ds12345_integrated.RData",
  compress = "gzip"
)
```

### Displaying Results

```{r}
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```

```{r}
covid19_ds124_analysis$volcano_plots
covid19_ds124_analysis$heatmap
covid19_ds124_analysis$ora_down_raw$enrichgo_visuals
#covid19_ds124_analysis$ora_up_raw$enrichgo_visuals
```

```{r}
heatmap_visuals_124

heatmap_folder <- "heatmap_images"


if(!dir.exists(heatmap_folder)) dir.create(heatmap_folder, recursive = TRUE)


for(h in seq_along(heatmap_visuals_124)){
gg_patchwork(heatmap_visuals_124[[h]], 
             glue("{heatmap_folder}/{names(heatmap_visuals_124)[h]}.png"),
             width = 10,
             height = 20)
}
```

# Meta-Analysis

### DEG Meta-Analysis

```{r}
library(ggvenn)
```

```{r}
deg_results_list <- list(
  ds1 = covid19_ds1_analysis$filtered_results,
  ds2 = covid19_ds2_analysis$filtered_results,
  ds4 = covid19_ds4_analysis$filtered_results,
  int_ds = covid19_ds124_analysis$filtered_results
)

# initialize output list
all_degs <- vector("list", length(deg_results_list))
names(all_degs) <- c("GSE202805", "GSE152641", "GSE227116", "Integrated")

# loop through each dataset
for (j in seq_along(deg_results_list)) {
  
  current_dataset <- deg_results_list[[j]]
  current_genes <- c()
  
  # loop through each DEG table in the dataset
  for (i in seq_along(current_dataset)) {
    
    current_genes <- c(
      current_genes,
      current_dataset[[i]]$geneid
    )
  }
  
  # optionally remove duplicates
  all_degs[[j]] <- unique(current_genes)
}
ggvenn(all_degs)
```

There is an overall large heterogeneous mixture of DEGs between the three individual data-sets and the batch corrected, integrated data-set. A majority of the DEGs reside within their parent data-set and remain exclusive to their internal comparisons. The Integrated Dataset does share at least 150 DEGs with GSE152641 (GSE152641 U Integrated nDEGs = 196), and GSE202805 (GSE202805 U Integrated = 153), where the three held 172 DEGs (GSE202805 U GSE152641 U Integrated nDEGs = 172). However the integrated dataset does not share many DEGs (nDEGs\<20) with GSE227116.

### OverRepresented Pathways Meta-Analysis

```{r}
ora_desc_collapse_fxn <- function(unpacked_results){
  ora_desc_list <- vector("list", length(unpacked_ora_results_ds1))
  names(ora_desc_list) <- names(unpacked_ora_results_ds1)
  
  for(i in seq_along(unpacked_ora_results_ds1)) {
    
    # for each comparison within the unpacked results
    current_comparison <- unpacked_ora_results_ds1[[i]]
    
    ora_descriptions <- c()
    
    # for each table within that specific comparison
    for(j in seq_along(current_comparison)) {
      
      ora_descriptions <- c(
        ora_descriptions,
        current_comparison[[j]]$gene_desc.x
      )
    }
    
    ora_desc_list[[i]] <- unique(ora_descriptions)
  }
  return(ora_desc_list)
}
```

```{r}
upreg_unpacked_ora_ds1 <- covid19_ds1_analysis$unpacked_results_up
downreg_unpacked_ora_ds1 <- covid19_ds1_analysis$unpacked_results_down

covid19_ds1_summarized_ora_terms <- list("up" = ora_desc_collapse_fxn(upreg_unpacked_ora_ds1),
                                   "down" = ora_desc_collapse_fxn(downreg_unpacked_ora_ds1)) %>% unlist()
                                   
                                  
upreg_unpacked_ora_ds2 <- covid19_ds2_analysis$unpacked_results_up
downreg_unpacked_ora_ds2 <- covid19_ds2_analysis$unpacked_results_down                      


covid19_ds2_summarized_ora_terms <- list("up" = ora_desc_collapse_fxn(upreg_unpacked_ora_ds2),
                                   "down" = ora_desc_collapse_fxn(downreg_unpacked_ora_ds2)) %>% unlist()


upreg_unpacked_ora_ds4 <- covid19_ds4_analysis$unpacked_results_up
downreg_unpacked_ora_ds4 <- covid19_ds4_analysis$unpacked_results_down                      


covid19_ds4_summarized_ora_terms <- list("up" = ora_desc_collapse_fxn(upreg_unpacked_ora_ds4),
                                   "down" = ora_desc_collapse_fxn(downreg_unpacked_ora_ds4)) %>% unlist()

upreg_unpacked_ora_ds124 <- covid19_ds124_analysis$unpacked_results_up
downreg_unpacked_ora_ds124 <- covid19_ds124_analysis$unpacked_results_down                      


covid19_ds124_summarized_ora_terms <- list("up" = ora_desc_collapse_fxn(upreg_unpacked_ora_ds124),
                                   "down" = ora_desc_collapse_fxn(downreg_unpacked_ora_ds124)) %>% unlist()
```

```{r}
all_ora_terms <- list("GSE202805" = covid19_ds1_summarized_ora_terms , 
                      "GSE152641" = covid19_ds2_summarized_ora_terms, 
                      "GSE227116" = covid19_ds4_summarized_ora_terms, 
                      "Integrated" = covid19_ds124_summarized_ora_terms)
ggvenn(all_ora_terms)
```

# DEGs from OverRepresented Pathways analysis Heatmap

# Healthy vs COVID-19 Comparisons

Covid-19 Dataset 1 (GSE202805) top over-represented pathways by p-values of hypergeometric test

```{r}
df1_ds1 <- covid19_ds1_analysis$unpacked_results_down$`healthy-covid19`$`immunoglobulin mediated immune response`
df2_ds1 <- covid19_ds1_analysis$unpacked_results_down$`healthy-covid19`$`B cell mediated immunity`
df3_ds1 <- covid19_ds1_analysis$unpacked_results_down$`healthy-covid19`$`antibacterial humoral response`
df4_ds1 <- covid19_ds1_analysis$unpacked_results_down$`healthy-covid19`$`mucosal immune response`
df5_ds1 <- covid19_ds1_analysis$unpacked_results_down$`healthy-covid19`$`antimicrobial humoral response`

df_ora_filtered <- unique(bind_rows(df1_ds1, df2_ds1, df3_ds1, df4_ds1, df5_ds1), by = "gene_id")


df_ora_filtered <- list(df_ora_filtered)
names(df_ora_filtered) <- 'healthy-covid19'


heatmap_function(dds_object, df_ora_filtered)
```

Covid-19 Dataset 2 (GSE152641) top over-represented pathways by p-values of hypergeometric test

```{r, include=FALSE}
df1_ds2 <- unpacked_results_down$`Healthy-COVID19`$`defense response to virus`
df1_ds2_ora_filtered <- unique(bind_rows(df1_ds2), by = "gene_id")

df1_ds2_ora_filtered <- list(df1_ds2_ora_filtered)
names(df1_ds2_ora_filtered) <- 'Healthy-COVID19'

image <- heatmap_function(dds_object, df1_ds2_ora_filtered)
```

```{r}
image
```

```{r}
table(colnames(expr)[matching_cols])
length(unique(colnames(expr)[matching_cols]))
ncol(expr)
```

# Healthy vs Convalescent Comparisons

(per Dataset)

# Covid-19 vs Convalescent Comparison

(per Dataset)
