---
title: "Project_1"
output:
  pdf_document: default
  html_document: default
---

Source Challenge: <https://github.com/mitreac/BIDS2026_projects>

```{r, include = FALSE}
library(tidyverse)
library(devtools)
library(GEOquery)
library(vroom)
library(biomaRt)
library(dplyr)
library(org.Hs.eg.db)
library(DESeq2)
library(clusterProfiler)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(RNAseqQC)
```

```{r}

```

## Reading in GSE202805 file

80 Samples:

Platform: Illumina HiSeq 4000 (Homo sapiens)

Case Groups: 4 acute-Mild, 6 Acute Moderate, 32 Acute-Severe, 28 Covalescent

Control Groups: 10 healthy controls

Total:

```{r}
gse_202805 <- getGEO("GSE202805", GSEMatrix = TRUE)

gse_202805 <- gse_202805[[1]]

expr_202805 <- exprs(gse_202805)

meta_202805 <- pData(gse_202805)

head(expr_202805)
head(meta_202805)

getGEOSuppFiles("GSE202805")
```

```{r}
path <- "./GSE202805/GSE202805_genecounts_20220406.txt.gz"

counts <- read.delim(gzfile(path))

colnames(counts) <- sub("\\.bam$", "", colnames(counts))

counts <- counts %>% dplyr::select(-c(Chr, Start, End, Strand, Length))
```

```{r}
meta_data <- meta_202805
meta_data
meta_data <- meta_data %>%
  rownames_to_column(var = "samples") %>%
  dplyr::select(samples, title, characteristics_ch1) %>%
  mutate(characteristics_ch1 = gsub("group:", "", characteristics_ch1)) %>%
  dplyr::select(-c(samples))

mapping <- c(meta_data$characteristics_ch1)
names(mapping) <- meta_data$title

colnames(counts) <- mapping[colnames(counts)] 

colnames(counts) <- paste(names(mapping), colnames(counts), sep = "_")

colnames(counts)[1] <- "geneid"


# Character Specific Clean Up
counts$geneid <- gsub("\\..*", "", counts$geneid)
colnames(counts) <- gsub(" ", "", colnames(counts))

# So this is a crucial part to remember lol is that acute-covid19, convalescant covid
#added waay late into the game after first round of harmonization and integration
colnames(counts) <- gsub("acute", "covid19", colnames(counts))

# MetaData construction
metadata <- data.frame(
  sample = colnames(counts),
  condition = dplyr::case_when(
    grepl("_covid19$", colnames(counts)) ~ "covid19",
    grepl("_convalescent$", colnames(counts)) ~ "convalescent",
    grepl("healthy$", colnames(counts)) ~ "healthy",
    TRUE ~ NA_character_
  ) 
) %>% slice(., 2:nrow(.)) %>%
  dplyr::mutate(condition = as.factor(condition)) %>%
  dplyr::mutate(sample = as.factor(sample))
```

```{r}
# summarizing replicated genes across all samples
counts <- counts %>%
  group_by(geneid) %>%
  summarise(across(where(is.numeric), sum))

counts <- counts %>% column_to_rownames(var = "geneid")


metadata[] <- lapply(metadata, factor)

```

```{r}
counts
metadata
```

```{r}
# Unflitered Counts Saved to a DeSeqDataSet (dds object)
covid19_ds1 <- make_dds(counts = counts , metadata = metadata, ah_record = "AH52260")
#renaming counts to raw counts
assayNames(covid19_ds1)[1] <- "counts"

#variance stabilized Raw counts
assays(covid19_ds1)[[2]] <- vst(covid19_ds1@assays@data$counts)
assayNames(covid19_ds1)[2] <- "var_stable"
#log_transformed (log normalized counts)
assays(covid19_ds1)[[3]] <- log10(covid19_ds1@assays@data$counts + 1)
assayNames(covid19_ds1)[3] <- "log_counts"
```

```{r}
colData(covid19_ds1)
```

```{r}
saveRDS(covid19_ds1, "./post_processed_data/gse202805_covid19_ds1.rds")
```

```{r}
plot_total_counts(covid19_ds1)
plot_library_complexity(covid19_ds1)
plot_gene_detection(covid19_ds1)
```

```{r}
head(rowData(covid19_ds1))
```

```{r}
plot_biotypes
plot_biotypes(covid19_ds1)
```

## Reading in GSE152641

65 blood Samples recoved form 46 COVID-19 convalescent patients and 10 healthy donors were analyzed by RNA- Sequencing

Platform:: NovaSeq 6000

10 healthd donors, 65 blood samples.

```{r}
gse_152641 <- getGEO("GSE152641", GSEMatrix = TRUE)
gse_152641 <- gse_152641[[1]]

expr_152641 <- exprs(gse_152641)
meta_152641 <- pData(gse_152641)

head(expr_152641)
head(meta_152641)

getGEOSuppFiles("GSE152641")
```

```{r}
meta_152641
```

```{r}
library(glue)
warned <- FALSE
acc_num <- "GSE152641"

url_ds2 <- "https://www.ncbi.nlm.nih.gov/geo/download/?format=file&type=rnaseq_counts"
path_ds2 <- paste(url_ds2, glue("acc={acc_num}"), glue("file={acc_num}_raw_counts_GRCh38.p13_NCBI.tsv.gz"), sep="&");



tbl_ds2 <- tryCatch(

  {
    as.matrix(data.table::fread(path_ds2, header=T, colClasses="integer"), rownames=1) %>% as.data.frame()
  },

  warning = function(w){

    message("Initial call unsuccessful. Using local GEO file.")

    path_ds2 <- list.files(acc_num,recursive = TRUE,full.names = TRUE)[1]

    as.matrix(data.table::fread(path_ds2, header=T, colClasses="integer"), rownames=1) %>% as.data.frame()
  }

)
```

```{r}
# cleaning the meta_data
meta_152641 <- meta_152641 %>%
            rename("condition" = "disease:ch1") %>%
            dplyr::mutate(condition = gsub("control", "", condition)) %>% 
            dplyr::mutate(condition = trimws(condition)) %>%
            dplyr::mutate(condition = gsub("s$","", condition)) %>%
            dplyr::mutate(condition = tolower(condition)) %>%
            dplyr::select(geo_accession,
                          "age:ch1", 
                          condition, 
                          "Sex:ch1")

colnames(meta_152641) <- sub(":.*", "", colnames(meta_152641))

meta_152641[] <- lapply(meta_152641, factor)

#Mapping sample to the Condition
mapping <- c(meta_152641$condition)
names(mapping) <- meta_152641$geo_accession
colnames(tbl_ds2) <- names(mapping)

new_names <- paste(colnames(tbl_ds2), mapping, sep = "_")
colnames(tbl_ds2) <- new_names
```

```{r}
meta_152641
```

```{r}

tbl_ds2 <- rownames_to_column(tbl_ds2, "geneid")

# Summarizing the repeated geneids
tbl_ds2 <- tbl_ds2 %>%
  group_by(geneid) %>%
  summarise(across(where(is.numeric), sum))

nrow(tbl_ds2)
# converting entrezids  to ensemblids


# Iterative conversion function
# arguments:input, from, to, iterations, species

#managig RAM Space on a 16Gb vector limit 
rm(expr_152641, gse_152641)
```

```{r}
tbl_ds2
meta_152641
```

# Total genes ids after summarization and concatentation of gene counts : 39,376

function layers

1.  mapIds from AnnotationDbi
2.  getBM from cluster Profiler
3.  bitr from clusterProfiler
4.  gconvert from g:Profiler

```{r}


# initial args
from_type <- "ENTREZ"

to_type <- "ENSEMBL"

org_db <- org.Hs.eg.db

# Harmoizing between each of the available annotation functions
harmonizer <- list(
    ENTREZ = list(local = "ENTREZID", biomart = "entrezgene_id", gconvert = "ENTREZGENE_ACC"),
    ENSEMBL = list(local = "ENSEMBL",  biomart = "ensembl_gene_id", gconvert = "ENSG"),
    SYMBOL  = list(local = "SYMBOL",   biomart = "external_gene_name",
                   gconvert = "SYMBOL")
  )

local_from   <- harmonizer[[from_type]]$local
local_to     <- harmonizer[[to_type]]$local
  
biomart_from <- harmonizer[[from_type]]$biomart
biomart_to   <- harmonizer[[to_type]]$biomart


# enter mapping
input_ids <- tbl_ds2$geneid

unique_ids <- unique(input_ids)


#Intialize final results
unique_ids_len <- length(unique_ids)

final_results <- c(replicate(unique_ids_len, NA))

names(final_results) <- unique_ids
```

# GeneID Conversions for Object

```{r}
#Converting EntrezIDs into Symbols
id_chunks <- split(unique_ids, ceiling(seq_along(unique_ids) / 10000))

for(i in seq_along(id_chunks)){
  chunk_map <- AnnotationDbi::mapIds(org_db, 
                      keys = id_chunks[[i]],
                      keytype = local_from,
                      column =   local_to,
                      multiVals = "first")
  #updating the inital final results
  final_results[names(chunk_map)] <- chunk_map
}

final_results <- unlist(final_results)
missing_ids <- names(final_results)[is.na(final_results)]
length(missing_ids)



if (length(missing_ids) > 0){
  
      layer2 <- getBM(
        attributes = c(biomart_from, biomart_to),
        filters = biomart_from,
        values = missing_ids,
        mart = useMart("ensembl", dataset = "hsapiens_gene_ensembl")
      )
      layer2
  if(nrow(layer2) > 0) {
    indices <- match(layer2[[biomart_from]], names(final_results))
    final_results[indices] <- layer2[[biomart_to]]
  }
}

missing_ids <- names(final_results)[is.na(final_results)]
length(missing_ids)
```

Is it possible to get the converted symbols into ensemble IDS in totality?

If this works then we can use multiple dbs on the symbols as a universal token

then convert them all back to Ensemble IDs

match their ENTREZID index back to the symbols then boom

we now have a totally converted list of ENTREZ –\> Symbol –\> ENSEMBL pipeline

Journal Note:

I'd rather have this project done and posted than trying to perfect the

Although it was a good try it's likely that those missing 1092X gene are ncRNA or tRNA and not protein coding GENEs

Important Note Down: Resulting dds object for bulk transcriptomic data of the second data set will not have all genes. 10,925 will be missing because the ENSEMBL IDS were not available.

If there is someone out there that can convert those remaining 10,925

```{r}
mapping_df <- data.frame(geneid = names(final_results),
                         ensembl_id = as.character(final_results))

final_tbl_ds2 <- tbl_ds2 %>% 
  left_join(., mapping_df, by = "geneid") %>%
  dplyr::relocate(ensembl_id, .after = "geneid")

#Slicing Rows with any NAs within the Data
final_tbl_ds2 <- final_tbl_ds2[complete.cases(final_tbl_ds2),]


final_tbl_ds2 <- final_tbl_ds2 %>%
    group_by(ensembl_id) %>%
    summarise(across(where(is.numeric), sum)) %>%
    column_to_rownames(var = "ensembl_id")

#final edits of the metadata
meta_152641$samples <- paste0(meta_152641$geo_accession, "_", meta_152641$condition)


rownames(meta_152641) <- meta_152641$samples


meta_152641 <- meta_152641 %>% dplyr::select(-c("samples"))
meta_152641
```

```{r}
final_tbl_ds2
```

```{r}
# Unflitered Counts Saved to a DeSeqDataSet (dds object)
covid19_ds2 <- make_dds(counts = final_tbl_ds2 , metadata = meta_152641, ah_record = "AH89426")
#renaming counts to raw counts
assayNames(covid19_ds2)[1] <- "counts"

#variance stabilized Raw counts
assays(covid19_ds2)[[2]] <- vst(covid19_ds2@assays@data$counts)
assayNames(covid19_ds2)[2] <- "var_stable"
#log_transformed (log normalized counts)
assays(covid19_ds2)[[3]] <- log10(covid19_ds2@assays@data$counts + 1)
assayNames(covid19_ds2)[3] <- "log_counts"
```

```{r}
dim(covid19_ds2@assays@data$counts)
```

```{r}
saveRDS(covid19_ds2, "./post_processed_data/gse152641_covid19_ds2.rds")
```

```{r}
plot_total_counts(covid19_ds2)
plot_library_complexity(covid19_ds2)
plot_gene_detection(covid19_ds2)
plot_biotypes(covid19_ds2)
```

# Reading in GSE161731

Illumina NovaSeq 6000 (Homo sapiens)

46 individuals with PCR positive, sympomatic COVID-19, 14 were sampled at multiple time points

(early ≤10 days, middle 11-21 days, late \>21 days)

tissue()

**Categorical severity of Healthy and COVID-19 patients**

59 = acute respiratory infection

17 = Influenza

20 = bacterial pneumonia

19 = healthy controls

```{r}
gse_161731 <- getGEO("GSE161731", GSEMatrix = TRUE)
gse_161731 <- gse_161731[[1]]

expr_161731 <- exprs(gse_161731)
meta_161731<- pData(gse_161731)

head(expr_161731)
head(meta_161731)

getGEOSuppFiles("GSE161731")


urld_ds3 <- "https://www.ncbi.nlm.nih.gov/geo/download/?format=file&type=rnaseq_counts"
path_ds3 <- paste(urld_ds3, "acc=GSE161731", "file=GSE161731_raw_counts_GRCh38.p13_NCBI.tsv.gz", sep="&");
tbl_ds3 <- as.data.frame(as.matrix(data.table::fread(path_ds3, header=T, colClasses="integer"), rownames=1))
```

**Pre-Cleaning Meta.Data**

In some parts this data there is a note that explicity tells the used to excise certain samples at the end of GSE1617116's preprocessing we will delete those rows

```{r}
bad_samples <- c(meta_161731$geo_accession[
  meta_161731$description %in% "excluded from analyses for poor data quality"
])
```

```{r}
bad_samples
```

|     |
|:----|
|     |

```{r}
meta_161731_edit <- meta_161731 %>%
  
  mutate(characteristics_ch1.1 = gsub(">|<", "", characteristics_ch1.1)) %>%
  mutate(characteristics_ch1.1 = gsub("age: ", "", characteristics_ch1.1)) %>%
  mutate(age = as.integer(characteristics_ch1.1)) %>%
  
  # Gender
  mutate(characteristics_ch1.2 = gsub("gender: ", "", characteristics_ch1.2)) %>%
  rename(gender = characteristics_ch1.2) %>%
  
  # Race
  mutate(characteristics_ch1.3 = gsub("race: ", "", characteristics_ch1.3)) %>%
  rename(race = characteristics_ch1.3) %>%
  
  # Cohort
  mutate(characteristics_ch1.4 = gsub("cohort: ", "", characteristics_ch1.4)) %>%
  rename(condition = characteristics_ch1.4) %>%
  
  # Hospitalized
  mutate(characteristics_ch1.6 = gsub("hospitalized: ", "", characteristics_ch1.6)) %>%
  rename(hospitalized = characteristics_ch1.6) %>%
  
  # Tissue
  mutate(characteristics_ch1.7 = gsub("tissue: ", "", characteristics_ch1.7)) %>%
  rename(tissue = characteristics_ch1.7) %>%
  
  mutate(characteristics_ch1.8 = gsub("batch: ", "", characteristics_ch1.8)) %>%
  rename(batch_num = characteristics_ch1.8) %>%
  
  #time collected, making it a longitudinal study
  mutate(characteristics_ch1.5 = gsub("time_since_onset: ", "", characteristics_ch1.5)) %>%
  rename(time_point = characteristics_ch1.5) %>%

  # We slice off three rows to be excluded from the data which is why the row count in this decreases from 198 to 195 samples
  
  # Batch (Converted to Integer)
  dplyr::select(geo_accession, gender, race, description, condition, batch_num, age, time_point) %>%
  
  mutate(
  condition = tolower(condition),
  condition = gsub("cov other", "covother", condition)
)
#convert all columns into factor type
meta_161731_edit[] <- lapply(meta_161731_edit, factor)


meta_161731_edit
```

```{r}
meta_161731_edit %>% 
  group_by(condition) %>% 
  summarise(count = n())
```

### Mapping Sample Name to Condition

```{r}
#Mapping sample to the Condition
mapping <- meta_161731_edit$condition
names(mapping) <- meta_161731_edit$geo_accession
new_names <- paste(colnames(tbl_ds3), mapping, sep = "_")
colnames(tbl_ds3) <- new_names
tbl_ds3 <- rownames_to_column(tbl_ds3, "geneid")
tbl_ds3
```

```{r}
tbl_ds3
meta_161731_edit
```

```{r}
# initial args
from_type <- "ENTREZ"

to_type <- "ENSEMBL"

org_db <- org.Hs.eg.db

# Harmoizing between each of the available annotation functions
harmonizer <- list(
    ENTREZ = list(local = "ENTREZID", biomart = "entrezgene_id", gconvert = "ENTREZGENE_ACC"),
    ENSEMBL = list(local = "ENSEMBL",  biomart = "ensembl_gene_id", gconvert = "ENSG"),
    SYMBOL  = list(local = "SYMBOL",   biomart = "external_gene_name",
                   gconvert = "SYMBOL")
  )

local_from   <- harmonizer[[from_type]]$local
local_to     <- harmonizer[[to_type]]$local
  
biomart_from <- harmonizer[[from_type]]$biomart
biomart_to   <- harmonizer[[to_type]]$biomart




unique_ids <- unique(tbl_ds3$geneid)


#Intialize final results
unique_ids_len <- length(unique_ids)

final_results <- c(replicate(unique_ids_len, NA))

names(final_results) <- unique_ids

#Converting EntrezIDs into Symbols
id_chunks <- split(unique_ids, ceiling(seq_along(unique_ids) / 10000))

for(i in seq_along(id_chunks)){
  chunk_map <- AnnotationDbi::mapIds(org_db, 
                      keys = id_chunks[[i]],
                      keytype = local_from,
                      column =   local_to,
                      multiVals = "first")
  #updating the inital final results
  final_results[names(chunk_map)] <- chunk_map
}



#Converting EntrezIDs into Symbols
id_chunks <- split(unique_ids, ceiling(seq_along(unique_ids) / 10000))

for(i in seq_along(id_chunks)){
  chunk_map <- AnnotationDbi::mapIds(org_db, 
                      keys = id_chunks[[i]],
                      keytype = local_from,
                      column =   local_to,
                      multiVals = "first")
  #updating the inital final results
  final_results[names(chunk_map)] <- chunk_map
}

final_results <- unlist(final_results)
missing_ids <- names(final_results)[is.na(final_results)]
length(missing_ids)



if (length(missing_ids) > 0){
  
      layer2 <- getBM(
        attributes = c(biomart_from, biomart_to),
        filters = biomart_from,
        values = missing_ids,
        mart = useMart("ensembl", dataset = "hsapiens_gene_ensembl")
      )
      layer2
  if(nrow(layer2) > 0) {
    indices <- match(layer2[[biomart_from]], names(final_results))
    final_results[indices] <- layer2[[biomart_to]]
  }
}

missing_ids <- names(final_results)[is.na(final_results)]
length(missing_ids)
```

```{r}
mapping_df <- data.frame(geneid = names(final_results),
                         ensembl_id = as.character(final_results))

final_tbl_ds3 <- tbl_ds3 %>% 
  left_join(., mapping_df, by = "geneid") %>%
  dplyr::relocate(ensembl_id, .after = "geneid")

#Slicing Rows with any NAs within the Data
final_tbl_ds3 <- final_tbl_ds3[complete.cases(final_tbl_ds3),]


final_tbl_ds3 <- final_tbl_ds3 %>%
    group_by(ensembl_id) %>%
    summarise(across(where(is.numeric), sum)) %>%
    column_to_rownames(var = "ensembl_id") %>%
    dplyr::select(-matches(bad_samples))
```

```{r}
#final meta_data editing
meta_161731_edit$samples <- paste0(meta_161731_edit$geo_accession, "_", meta_161731_edit$condition)
meta_161731_edit <- meta_161731_edit %>% 
  dplyr::filter(description != 'excluded from analyses for poor data quality') %>%
  relocate(samples, .after = geo_accession) %>% 
  dplyr::select(samples, gender, race, condition, batch_num, age) %>%
  separate(samples, 
           into = c("sample_id", "condition"),
           sep = "_")
rownames(meta_161731_edit) <- NULL


#final edits of the metadata
meta_161731_edit$samples <- paste0(meta_161731_edit$sample_id, "_", meta_161731_edit$condition)

meta_161731_edit[] <- lapply(meta_161731_edit, factor)

rownames(meta_161731_edit) <- meta_161731_edit$samples
```

```{r}
meta_161731_edit
final_tbl_ds3
```

```{r}
# Unflitered Counts Saved to a DeSeqDataSet (dds object)
covid19_ds3 <- make_dds(counts = final_tbl_ds3 , metadata = meta_161731_edit, ah_record = "AH89426")
#renaming counts to raw counts
assayNames(covid19_ds3)[1] <- "counts"

#variance stabilized Raw counts
assays(covid19_ds3)[[2]] <- vst(covid19_ds3@assays@data$counts)
assayNames(covid19_ds3)[2] <- "var_stable"

#log_transformed (log normalized counts)
assays(covid19_ds3)[[3]] <- log10(covid19_ds3@assays@data$counts + 1)
assayNames(covid19_ds3)[3] <- "log_counts"
```

```{r}
saveRDS(covid19_ds3, "./post_processed_data/gse161731_covid19_ds3.rds")
```

```{r}
missing_samples <- rownames(meta_161731_edit)[
  !(rownames(meta_161731_edit) %in% colnames(final_tbl_ds3))
]
missing_samples
```

# Reading in GSE227116

```{r}
gse_227116 <- getGEO("GSE227116", GSEMatrix = TRUE)

gse_227116 <- gse_227116[[1]]

expr_227116 <- exprs(gse_227116)

meta_227116 <- pData(gse_227116)

head(expr_227116)
head(meta_227116)

getGEOSuppFiles("GSE227116")
```

```{r}
library(glue)
warned <- FALSE
acc_num <- "GSE227116"

url_ds4 <- "https://www.ncbi.nlm.nih.gov/geo/download/?format=file&type=rnaseq_counts"

path_ds4 <- paste(url_ds4, glue("acc={acc_num}"), glue("file={acc_num}_raw_counts_GRCh38.p13_NCBI.tsv.gz"), sep="&");



tbl_ds4 <- tryCatch(

  {
    as.matrix(data.table::fread(path_ds4, header=T, colClasses="integer"), rownames=1) %>% as.data.frame()
  },

  warning = function(w){

    message("Initial call unsuccessful. Using local GEO file.")

    path_ds4 <- list.files(acc_num , recursive = TRUE , full.names = TRUE)[1]

    as.matrix(data.table::fread(path_ds4, header=T, colClasses="integer"), rownames=1) %>% as.data.frame()
  }

)
```

```{r}
meta_227116
```

```{r}
meta_227116_edit <- meta_227116 %>% 
  rename(condition = `status:ch1`) %>% 
  mutate(
    condition = tolower(condition),
    condition = gsub("covid-19", "", condition),
    condition = trimws(condition),
    condition = gsub("s$", "", condition),
    conition = tolower(condition)
  ) %>%
  mutate(samples = paste0(geo_accession, "_", condition)) %>% 
  dplyr::select(geo_accession, samples, "time point:ch1", condition) %>%
  rename(time_point = "time point:ch1")
rownames(meta_227116_edit) <- meta_227116_edit$samples


meta_227116_edit[] <- lapply(meta_227116_edit, factor)

```

```{r}
mapping <- rownames(meta_227116_edit)
names(mapping) <- colnames(tbl_ds4)
colnames(tbl_ds4) <- mapping
tbl_ds4 <- rownames_to_column(tbl_ds4,"geneid")
```

```{r}
# initial args
from_type <- "ENTREZ"

to_type <- "ENSEMBL"

org_db <- org.Hs.eg.db

# Harmoizing between each of the available annotation functions
harmonizer <- list(
    ENTREZ = list(local = "ENTREZID", biomart = "entrezgene_id", gconvert = "ENTREZGENE_ACC"),
    ENSEMBL = list(local = "ENSEMBL",  biomart = "ensembl_gene_id", gconvert = "ENSG"),
    SYMBOL  = list(local = "SYMBOL",   biomart = "external_gene_name",
                   gconvert = "SYMBOL")
  )

local_from   <- harmonizer[[from_type]]$local
local_to     <- harmonizer[[to_type]]$local
  
biomart_from <- harmonizer[[from_type]]$biomart
biomart_to   <- harmonizer[[to_type]]$biomart




unique_ids <- unique(tbl_ds4$geneid)


#Intialize final results
unique_ids_len <- length(unique_ids)

final_results <- c(replicate(unique_ids_len, NA))

names(final_results) <- unique_ids

#Converting EntrezIDs into Symbols
id_chunks <- split(unique_ids, ceiling(seq_along(unique_ids) / 10000))

for(i in seq_along(id_chunks)){
  chunk_map <- AnnotationDbi::mapIds(org_db, 
                      keys = id_chunks[[i]],
                      keytype = local_from,
                      column =   local_to,
                      multiVals = "first")
  #updating the inital final results
  final_results[names(chunk_map)] <- chunk_map
}



#Converting EntrezIDs into Symbols
id_chunks <- split(unique_ids, ceiling(seq_along(unique_ids) / 10000))

for(i in seq_along(id_chunks)){
  chunk_map <- AnnotationDbi::mapIds(org_db, 
                      keys = id_chunks[[i]],
                      keytype = local_from,
                      column =   local_to,
                      multiVals = "first")
  #updating the inital final results
  final_results[names(chunk_map)] <- chunk_map
}

final_results <- unlist(final_results)
missing_ids <- names(final_results)[is.na(final_results)]
length(missing_ids)



if (length(missing_ids) > 0){
  
      layer2 <- getBM(
        attributes = c(biomart_from, biomart_to),
        filters = biomart_from,
        values = missing_ids,
        mart = useMart("ensembl", dataset = "hsapiens_gene_ensembl")
      )
      layer2
  if(nrow(layer2) > 0) {
    indices <- match(layer2[[biomart_from]], names(final_results))
    final_results[indices] <- layer2[[biomart_to]]
  }
}

missing_ids <- names(final_results)[is.na(final_results)]
length(missing_ids)

mapping_df <- data.frame(geneid = names(final_results),
                         ensembl_id = as.character(final_results))
```

```{r}
final_tbl_ds4 <- tbl_ds4 %>% 
  left_join(., mapping_df, by = "geneid") %>%
  dplyr::relocate(ensembl_id, .after = "geneid")

#Slicing Rows with any NAs within the Data
final_tbl_ds4 <- final_tbl_ds4[complete.cases(final_tbl_ds4),]


final_tbl_ds4 <- final_tbl_ds4 %>%
    group_by(ensembl_id) %>%
    summarise(across(where(is.numeric), sum)) %>%
    column_to_rownames(var = "ensembl_id")
```

```{r}
meta_227116_edit
final_tbl_ds4
```

```{r}
meta_227116_edit <- meta_227116_edit %>% 
  relocate(samples, .after = geo_accession) %>% 
  separate(samples, 
           into = c("sample_id", "condition"),
           sep = "_")
```

```{r}
covid19_ds4 <- make_dds(final_tbl_ds4, metadata = meta_227116_edit, ah_record = "AH89426")
#renaming counts to raw counts
assayNames(covid19_ds4)[1] <- "counts"

#variance stabilized Raw counts
assays(covid19_ds4)[[2]] <- vst(covid19_ds4@assays@data$counts)
assayNames(covid19_ds4)[2] <- "var_stable"

#log_transformed (log normalized counts)
assays(covid19_ds4)[[3]] <- log10(covid19_ds4@assays@data$counts + 1)
assayNames(covid19_ds4)[3] <- "log_counts"
```

```{r}
saveRDS(covid19_ds4, "./post_processed_data/gse227116_covid19_ds4.rds")
```

# Reading in GSE152418

Illumina NovaSeq 6000

PBMC

17 COVID-19, Healthy Controls-17

Total Samples = 34

```{r}
gse_152418 <- getGEO("GSE152418", GSEMatrix = TRUE)

gse_152418 <- gse_152418[[1]]

expr_152418 <- exprs(gse_152418)

meta_152418 <- pData(gse_152418)

head(expr_152418)
head(meta_152418)

getGEOSuppFiles("GSE152418")
```

```{r}
library(glue)
warned <- FALSE
acc_num <- "GSE152418"

url_ds5 <- "https://www.ncbi.nlm.nih.gov/geo/download/?format=file&type=rnaseq_counts"

path_ds5 <- paste(url_ds5, glue("acc={acc_num}"), glue("file={acc_num}_raw_counts_GRCh38.p13_NCBI.tsv.gz"), sep="&");



tbl_ds5 <- tryCatch(

  {
    as.matrix(data.table::fread(path_ds5, header=T, colClasses="integer"), rownames=1) %>% as.data.frame()
  },

  warning = function(w){

    message("Initial call unsuccessful. Using local GEO file.")

    path_ds5 <- list.files(acc_num , recursive = TRUE , full.names = TRUE)[1]

    as.matrix(data.table::fread(path_ds5, header=T, colClasses="integer"), rownames=1) %>% as.data.frame()
  }

)
```

```{r}
meta_152418
tbl_ds5
```

Meta_Data Editing

```{r}
meta_152418_edit <- meta_152418 %>%
  rename(condition = 'disease state:ch1') %>% 
  rename(days_post_symptom_onset = `days_post_symptom_onset:ch1`) %>%
  rename(severity = `severity:ch1`) %>%
  mutate(
    condition = tolower(condition),
    condition = trimws(condition),
    condition = gsub("s$", "", condition),
    condition = tolower(condition),
    condition = gsub("covid-19", "covid19", condition)
  ) %>%
  mutate(samples = paste0(geo_accession, "_", condition)) %>% 
  dplyr::select(geo_accession, samples, condition, `days_post_symptom_onset`
, `severity`)

meta_152418[] <- lapply(meta_152418, factor)

rownames(meta_152418_edit) <- NULL

meta_152418_edit <- meta_152418_edit %>%
  column_to_rownames("samples")


mapping <- rownames(meta_152418_edit)
names(mapping) <- colnames(tbl_ds5)
colnames(tbl_ds5) <- mapping
tbl_ds5 <- rownames_to_column(tbl_ds5,"geneid")
```

```{r}
meta_152418_edit
```

```{r}
# initial args
from_type <- "ENTREZ"

to_type <- "ENSEMBL"

org_db <- org.Hs.eg.db

# Harmoizing between each of the available annotation functions
harmonizer <- list(
    ENTREZ = list(local = "ENTREZID", biomart = "entrezgene_id", gconvert = "ENTREZGENE_ACC"),
    ENSEMBL = list(local = "ENSEMBL",  biomart = "ensembl_gene_id", gconvert = "ENSG"),
    SYMBOL  = list(local = "SYMBOL",   biomart = "external_gene_name",
                   gconvert = "SYMBOL")
  )

local_from   <- harmonizer[[from_type]]$local
local_to     <- harmonizer[[to_type]]$local
  
biomart_from <- harmonizer[[from_type]]$biomart
biomart_to   <- harmonizer[[to_type]]$biomart




unique_ids <- unique(tbl_ds5$geneid)


#Intialize final results
unique_ids_len <- length(unique_ids)

final_results <- c(replicate(unique_ids_len, NA))

names(final_results) <- unique_ids

#Converting EntrezIDs into Symbols
id_chunks <- split(unique_ids, ceiling(seq_along(unique_ids) / 10000))

for(i in seq_along(id_chunks)){
  chunk_map <- AnnotationDbi::mapIds(org_db, 
                      keys = id_chunks[[i]],
                      keytype = local_from,
                      column =   local_to,
                      multiVals = "first")
  #updating the inital final results
  final_results[names(chunk_map)] <- chunk_map
}



#Converting EntrezIDs into Symbols
id_chunks <- split(unique_ids, ceiling(seq_along(unique_ids) / 10000))

for(i in seq_along(id_chunks)){
  chunk_map <- AnnotationDbi::mapIds(org_db, 
                      keys = id_chunks[[i]],
                      keytype = local_from,
                      column =   local_to,
                      multiVals = "first")
  #updating the inital final results
  final_results[names(chunk_map)] <- chunk_map
}

final_results <- unlist(final_results)
missing_ids <- names(final_results)[is.na(final_results)]
length(missing_ids)



if (length(missing_ids) > 0){
  
      layer2 <- getBM(
        attributes = c(biomart_from, biomart_to),
        filters = biomart_from,
        values = missing_ids,
        mart = useMart("ensembl", dataset = "hsapiens_gene_ensembl")
      )
      layer2
  if(nrow(layer2) > 0) {
    indices <- match(layer2[[biomart_from]], names(final_results))
    final_results[indices] <- layer2[[biomart_to]]
  }
}

missing_ids <- names(final_results)[is.na(final_results)]
length(missing_ids)
```

```{r}
mapping_df <- data.frame(geneid = names(final_results),
                         ensembl_id = as.character(final_results))

final_tbl_ds5 <- tbl_ds5 %>% 
  left_join(., mapping_df, by = "geneid") %>%
  dplyr::relocate(ensembl_id, .after = "geneid")

#Slicing Rows with any NAs within the Data
final_tbl_ds5 <- final_tbl_ds5[complete.cases(final_tbl_ds5),]


final_tbl_ds5 <- final_tbl_ds5 %>%
    group_by(ensembl_id) %>%
    summarise(across(where(is.numeric), sum)) %>%
    column_to_rownames(var = "ensembl_id")
```

```{r}
covid19_ds5 <- make_dds(final_tbl_ds5, metadata = meta_152418_edit, ah_record = "AH89426")
#renaming counts to raw counts
assayNames(covid19_ds5)[1] <- "counts"

#variance stabilized Raw counts
assays(covid19_ds5)[[2]] <- vst(covid19_ds5@assays@data$counts)
assayNames(covid19_ds5)[2] <- "var_stable"

#log_transformed (log normalized counts)
assays(covid19_ds5)[[3]] <- log10(covid19_ds5@assays@data$counts + 1)
assayNames(covid19_ds5)[3] <- "log_counts"
```

```{r}
saveRDS(covid19_ds5, "./post_processed_data/gse152418_covid19_ds5.rds")
```

```{r}
missing_samples <- rownames(meta_161731_edit)[
  !(rownames(meta_152418_edit) %in% colnames(final_tbl_ds5))
]
missing_samples

missing_samples <- rownames(meta_152418_edit)[
  !(rownames(meta_152418_edit) %in% colnames(final_tbl_ds5))
]
missing_samples


```
