---
title: "Project_2_Data_Curation"
output: html_document
---

```{r,include =FALSE}
library(tidyverse)
library(devtools)
library(vroom)
library(GEOquery)
library(biomaRt)
library(dplyr)
library(org.Hs.eg.db)
library(DESeq2)
library(clusterProfiler)
library(RNAseqQC)
#library(bulkanalysis) # fix this later
#source("Project_1_function_lab.R") # fix this later
```

# Querying for RNA-Seq Dataset

### Acquiring Chosen IDs and Filtering for Covariates

Checking meta data for key features

Gleason Score

**Pathologic T Stage:** grades how far the cancer has anatomically spreads within a or beyond the prostate.Pathalogic T stage (pT2 vs. pT3/T4)

PSA level at diagnosis (Common threshold: \<10, 10–20, \>20 ng/mL)

Biochemical Recurrance status (Available as a binary label)

Source of Pathology/Molecular Report: <https://pmc.ncbi.nlm.nih.gov/articles/PMC4695400/#S14>

ajcc_pathologi_t

paper_PSA_preop:

Reviewed_Gleason_sum :

ETV1_status

ETV4_status

FOXA1_mut

Mutations

Fraction_genome_altered

AR_score

# Need justification for each of following Chosen Covariates.

Why would they be meaningful feature that sample embeddings cluster around?

```{r}
library(openxlsx)
pathology_path <- "./GDCdata/TCGA_PRAD_pathology.xlsx"
pathology_metadata <- read.xlsx(pathology_path)
pathology_metadata
```

```{r}
#pathology covariates
path_cov <- c("SAMPLE_ID", 
              "PATIENT_ID", 
              "Age", 
              "Race", 
              "Subtype",
              "Reviewed_Gleason_sum", 
              "ETV1_status", 
              "ETV4_status", 
              "FOXA1_mut", 
              "Mutations", 
              "Fraction_genome_altered", 
              "AR_score",
              "PSA_preop")

covariates <- c("ajcc_pathologic_t", "gleason_score", "paper_PSA_preop", "paper_AR_score","paper_FOXA1_mut", "paper_ETV1_status", "paper_ETV4_status")

updated_covariates <- c("ajcc_pathologic_t", 
                        "gleason_score", 
                        "paper_PSA_preop", 
                        "paper_AR_score",
                        "paper_FOXA1_mut", 
                        "paper_ETV1_status", 
                        "paper_ETV4_status",
                        "Mutations",
                        "Fraction_genome_altered")
```

```{r}
# ==========================
# Clinical covariates
# ==========================

clinical_covariates <- c(
  "age_at_diagnosis",
  "cases",
  "race",
  "gender",
  "paper_PSA_preop"
)

# ==========================
# Pathological covariates
# ==========================

pathological_covariates <- c(
  "gleason_score",
  "ajcc_pathologic_t",
  "ajcc_pathologic_n",
  "primary_diagnosis",
  "paper_Residual_tumor",
  "Subtype"
)

# ==========================
# Molecular covariates
# ==========================

molecular_covariates <- c(
  "paper_iCluster",
  "paper_mRNA_cluster",
  "paper_methylation_cluster",

  "paper_ERG_status",
  "paper_ETV1_status",
  "paper_ETV4_status",
  "paper_SPINK1_high",

  "paper_FOXA1_mut",
  "paper_SPOP_mut",
  "paper_TP53_mut",
  "paper_PTEN_mut",

  "paper_Mutations",
  "paper_Fraction_genome_altered",

  "paper_ABSOLUTE_purity",
  "paper_AR_score"
)

# ==========================
# Survival variables
# ==========================

survival_covariates <- c(
  "vital_status",
  "days_to_last_follow_up",
  "cause_of_death"
)

# ==========================
# Identifiers
# ==========================

id_covariates <- c(
  "PATIENT_ID",
  "SAMPLE_ID"
)

# ==========================
# Complete covariate list
# ==========================

all_covariates <- c(
  id_covariates,
  clinical_covariates,
  pathological_covariates,
  molecular_covariates,
  survival_covariates
)
```

Pulling query results for the TCGA-PRAD datasets

```{r}
library(TCGAbiolinks)

#grab the results query
#I need the actual data
query <- GDCquery(
  project = "TCGA-PRAD",
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
)
# Primary Tumor: Direct tumor samples
#Solid Tissue Normal: These patient samples are called "solid tissue normals" and are taken from normal tissues near the tumor
results <- getResults(query)
results
# FIX: extract submitter ID properly
colnames(results)[colnames(results) == "cases.submitter_id"] <- "PATIENT_ID"

# ensure pathology metadata is filtered correctly
pathology_metadata <- pathology_metadata %>%
  dplyr::select(any_of(path_cov))

pathology_metadata
# join
merged_meta <- results %>%
  left_join(pathology_metadata, by = "PATIENT_ID")

merged_meta
```

```{r}
View(results)
```

```{r}
#Using 
clean_df <- merged_meta[complete.cases(merged_meta [, c("PATIENT_ID", "Reviewed_Gleason_sum", "PSA_preop")]), ]


# Ending Df samples into 15
#still results of the results query and not the meta_data the SE object
clean_df_sample <- clean_df
clean_df_sample

#Download the 15 samples instead of packing in 554 different DataSets
barcodes <- clean_df_sample$cases

query_uuid <- GDCquery(
        project = "TCGA-PRAD",
        data.category = "Transcriptome Profiling",
        data.type = "Gene Expression Quantification",
        barcode = barcodes)
  
  
GDCdownload(query_uuid)
  
se <- GDCprepare(query_uuid)

#pulling the metadata from the downloaded tcga datasets
tcga_coldata <- as.data.frame(colData(se))
tcga_coldata


#rename the tcga_coldat colums to mere with pathology meta_data
colnames(tcga_coldata)[colnames(tcga_coldata) == "barcode"] <- "cases"

merged_coldata <- left_join(clean_df_sample, tcga_coldata, by = "cases")
```

```{r}
View(results)
```

```{r}
#Using 
clean_df <- merged_meta[complete.cases(merged_meta [, c("PATIENT_ID", "Reviewed_Gleason_sum", "PSA_preop")]), ]


# Ending Df samples into 15
#still results of the results query and not the meta_data the SE object
clean_df_sample <- clean_df
clean_df_sample

#Download the 15 samples instead of packing in 554 different DataSets
barcodes <- clean_df_sample$cases

query_uuid <- GDCquery(
        project = "TCGA-PRAD",
        data.category = "Transcriptome Profiling",
        data.type = "Gene Expression Quantification",
        barcode = barcodes)
  
  
GDCdownload(query_uuid)
  
se <- GDCprepare(query_uuid)

#pulling the metadata from the downloaded tcga datasets
tcga_coldata <- as.data.frame(colData(se))
tcga_coldata


#rename the tcga_coldat colums to mere with pathology meta_data
colnames(tcga_coldata)[colnames(tcga_coldata) == "barcode"] <- "cases"

merged_coldata <- left_join(clean_df_sample, tcga_coldata, by = "cases")
```

### Altering for Continous and Discrete Variables

```{r}
#Group Continuous Variables
alt_coldata$age_group <- cut(
  alt_coldata$age_at_diagnosis,
  breaks = c(0, 50, 60, 70, 80, Inf),
  labels = c("<50", "50-60", "60-70", "70-80", "80+"),
  include.lowest = TRUE
)

alt_coldata$PSA_group <- cut(
  alt_coldata$paper_PSA_preop,
  breaks = c(0, 5, 10, 15, 20, Inf),
  labels = c("<5", "5-10", "10-15", "15-20", "20+"),
  include.lowest = TRUE
)

alt_coldata$AR_score_group <- cut(
  alt_coldata$paper_AR_score,
  breaks = c(-20, -10 ,-5 ,0, 25, 50, 75, 100, Inf),
  labels = c("-20 - -10","-10 - -5","-5 - 0"," 0-25", "25-50", "50-75", "75-100", "100+"),
  include.lowest = TRUE
)

alt_coldata$ABSOLUTE_purity_group <- cut(
  alt_coldata$paper_ABSOLUTE_purity,
  breaks = c(0, 0.50, 0.70, 0.85, 1.00),
  include.lowest = TRUE,
  labels = c("Low", "Moderate", "High", "Very High")
)

alt_coldata$FGA_group <- cut(
  alt_coldata$paper_Fraction_genome_altered,
  breaks = c(0, 0.25, 0.50, 0.75, 1.00),
  include.lowest = TRUE,
  labels = c("0-0.25", "0.25-0.50", "0.50-0.75", "0.75-1.00")
)



alt_coldata
colData(se) <- DataFrame(alt_coldata)
```

```{r}
View(alt_coldata)
```

TCGA-PRAD: Project dataformat

m x n matrix where is a column wise samples. m = 60k genes, and n = \# of barcodes/patient/samples requested from the database.

Essentially will be classfiying on the column basis instead of the dataset basis makes sense. I am used to seeing it as an mxn problem not an columun wise column.

Goal is classify each of these into a clinical label.

**Input:** TCGA Transcriptomic data Column, and the x amount of variable included for each of those columns of count data

**Output:** A clinical Label classification based on those effects

**Full Project Question:** Do molecular subtypes defined from transcriptomic data align with clincal classification?

How does molecular sub-typing of PRAD relate to the clinical classification ? How does transcriptomic data it help more or better than the clinical variables that are collected?

What makes PRAD soooo clinically heterogenous: "patients with similar histological grade can have vastly different outcomes"

How does molecular subtyping of PRAD relate to the clinical classification or at least inform clinical classification?

My job:

1.  Perform exploratory dim reduction an clustering for transcriptomic subtypes
2.  Use those as labels for the data per column/patient/sample
3.  Build the classifier based on the transcriptomic subtypes

## AnnData Conversion

```{r}
library(convert2anndata)
library(Seurat)
```

```{r}
#transfer the rownames of the sample onto the unstranded matrix
for (i in seq_along(assays(se))) {
  
  mat <- assays(se, withDimnames=FALSE)[[i]]
  
  colnames(mat) <- rownames(colData(se))
  rownames(mat) <- rownames(as.data.frame(rowRanges(se)))

  assays(se)[[i]] <- mat
  
}

#Okay learned something new. DDS objects don't accept TPM and Seurat also hates TPM normalized counts.
# This is because it's going off of library size expression and not direct relative expression
# Intersting Article about Seurat basics: https://github.com/satijalab/seurat/issues/7496
```

### Converting SE into h5ad

```{r}
library(convert2anndata)
library(anndata)
```

```{r, include = FALSE}
ad_path <- "./GDCdata/"

seurat_se <-  CreateSeuratObject(
  counts = assays(se)[["unstranded"]],
  meta.data = colData(se)
  
)
#Log noralizing unstranded data
seurat_se <- NormalizeData(seurat_se)
seurat_se <- FindVariableFeatures(seurat_se, nfeatures = 5000)
seurat_se <- ScaleData(seurat_se)
seurat_se <- RunPCA(seurat_se)
sce <- convert_seurat_to_sce(seurat_se)
ad <- convert_to_anndata(sce, assayName = "counts", useAltExp = TRUE)
write_h5ad(ad, paste0(ad_path, "seurat_se.h5ad"))


#Unfortunetly I need to create a seperate Seurat Assay
counts <- assays(se)[["tpm_unstrand"]]
seurat_se_tpm <- CreateSeuratObject(
  counts = counts,
  meta.data = as.data.frame(colData(se))
)
seurat_se_tpm <- NormalizeData(seurat_se_tpm, normalization.method = "LogNormalize")
seurat_se_tpm <- FindVariableFeatures(seurat_se_tpm, nfeatures = 5000)
seurat_se_tpm <- ScaleData(seurat_se_tpm)
seurat_se_tpm <- RunPCA(seurat_se_tpm)

sce_tpm  <- convert_seurat_to_sce(seurat_se_tpm)

ad_tpm <- convert_to_anndata(sce_tpm, assayName = "counts", useAltExp = TRUE)

write_h5ad(ad_tpm, paste0(ad_path, "seurat_se_tpm.h5ad"))
```

```{r}
#saving Curated Summarized Experiment
save_data <- "saved_data"
if(!dir.exists(save_data)) dir.create(save_data, recursive = TRUE)
saveRDS(se, "./saved_data/curated_RSE_object.rds")
```

**END OF PRE-PROCESSING NOTEBOOK**
