---
title: "Project_2_DimReduction"
output: html_document
---

Library Read In

```{r,include =FALSE}
library(tidyverse)
library(devtools)
library(vroom)
library(GEOquery)
library(biomaRt)
library(dplyr)
library(org.Hs.eg.db)
library(DESeq2)
library(clusterProfiler)
library(RNAseqQC)
#library(bulkanalysis) # fix this later
#source("Project_1_function_lab.R") # fix this later
```

# **DimReduction TPM + Log2 Normalization + Metadata coloring**

Reading in curated data file

```{r}
se <- readRDS("./saved_data/curated_RSE_object.rds")
```

```{r}
met <- as.data.frame(colData(se))
rownames <- rownames(as.data.frame(colData(se)))
met$SAMPLE_ID <- rownames
rownames(met) <- rownames(NULL)

met
```

```{r}
tpm.mat <- se@assays@data$tpm_unstrand

#TPM is a very left skewed measure we have to log scale it. 
log_tpm <- log2(tpm.mat + 1)


#Select top variable gene
top_var_genes <- head(order(rowVars(log_tpm), decreasing = TRUE), 5000)
vst_top <- log_tpm[top_var_genes, ]
# Run PCA via R
# Remember PCA is taking the right unitary martrix (VT) (feature space) of SVD
# VT tells us the principal directions of the data set
# multiply the the original matrix by the right unitary which is equivalent to the
# U time signma (left unitary) * Singular values (Size of the PCs)


#PCA on genes
pca_res <- prcomp(
  t(log_tpm[top_var_genes, ]),
  center = TRUE,
  scale. = TRUE)

pca_df <- data.frame(
  SAMPLE_ID = colnames(vst_top),
  PC1= pca_res$x[ ,1],
  PC2 = pca_res$x[ ,2]
)
#got a duplicate error. Don't know why R decided it's a problem now but had to make a workaround
met <- as.data.frame(colData(se))
rownames <- rownames(as.data.frame(colData(se)))
met$SAMPLE_ID <- rownames
rownames(met) <- rownames(NULL)
pca_df <- left_join(pca_df, met , by = "SAMPLE_ID")
```

###Scree Plot Making the Scree Plot to Verify Variance Ratio per PC

```{r}

vars <- colnames(pca_df)[5:ncol(pca_df)]
PCA_plots <- list()
for (v in vars) {

  p <- ggplot(
    pca_df,
    aes(PC1, PC2, color = factor(.data[[v]]))
  ) +
    geom_point() +
    ggtitle(glue::glue("PCA space with Var: {v}"))

  PCA_plots[[v]] <- p
}


 # Calculating Percentage of Variance Per PC for the Scree Plot
  
  # Calculate percentage variance explained
  var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100
  
  # Create a data frame for plotting (first 10 PCs)
  scree_data <- data.frame(
    PC = paste0("PC", 1:10),
    Variance = var_explained[1:10]
  )
  scree_data
  # Ensure PC is treated as a factor in order
  scree_data$PC <- factor(scree_data$PC, levels = scree_data$PC)
  
  # Scree Plot ensures that we are plotting PCs that capture the most amount of variability within our data
  scree_plot <- ggplot(scree_data, aes(x = PC, y = Variance, group = 1)) +
    geom_line(color = "red", size = 1) +
    geom_point(color = "red", size = 2) +
    labs(title = "Scree Plot: Variance Explained per PC",
         x = "Principal Component",
         y = "Percentage of Variance (%)") +
    theme_minimal()

scree_plot
```

PLotting Possibly Highly Stratified MetaVars onto the Scaled Left Unitary Matrix (U \* Sigma)

```{r}

#Clinical Covariates
PCA_plots[["PSA_group"]]
PCA_plots[["gleason_score"]]
PCA_plots[["ajcc_pathologic_t"]]
PCA_plots[["ajcc_pathologic_n"]]
PCA_plots$Subtype

#molecular covariates
PCA_plots[["paper_ETV1_status"]]
PCA_plots[["paper_FOXA1_mut"]]
PCA_plots[["FGA_group"]]
PCA_plots$paper_mRNA_cluster
PCA_plots$paper_iCluster
PCA_plots$AR_score_group
PCA_plots[["FGA_group"]]
PCA_plots$paper_ETV1_status
PCA_plots$paper_TP53_mut
```

# Testing Unstranded Counts + VST + PCA

```{r, include = FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
ah[["AH89426"]]
```

```{r, include = FALSE}
# Unflitered Counts Saved to a DeSeqDataSet (dds object)
tcgna_prad <- make_dds(counts = se@assays@data$unstranded , metadata = colData(se), ah_record = "AH89426")
#renaming counts to raw counts
assayNames(tcgna_prad)[1] <- "counts"

#variance stabilized Raw counts
assays(tcgna_prad)[[2]] <- vst(tcgna_prad@assays@data$counts)
assayNames(tcgna_prad)[2] <- "var_stable"
#log_transformed (log normalized counts)
assays(tcgna_prad)[[3]] <- log10(tcgna_prad@assays@data$counts + 1)
assayNames(tcgna_prad)[3] <- "log_counts"
```

```{r}
bulk_dataset <- tcgna_prad
anno_variable <- "gleason_score"
color_variable <-  "gleason_score"
#prioritize variance stable assay within the bulk_dataset
  assays(bulk_dataset) <- assays(bulk_dataset)[c("var_stable", "counts", "log_counts")]
  # takes the euclidan distance of a variance stabilized dataset
  # row by row euclidean distance against each other
  # plots the visual distance between each sample onto heatmap
  distance_plot <- plot_sample_clustering(bulk_dataset, anno_vars = anno_variable, distance = "euclidean")


  #____Extracting Top Variable Gene per sample -----#
  # Setting Variance Stable Counts as Priority Assay
  vst_mat <- assay(bulk_dataset, "var_stable")
  #Select top variable gene
  top_var_genes <- head(order(rowVars(vst_mat), decreasing = TRUE), 2000)
  vst_top <- vst_mat[top_var_genes, ]
  # Run PCA via R
  # Remember PCA is taking the right unitary martrix (VT) (feature space) of SVD
  # VT tells us the principal directions of the data set
  # multiply the the original matrix by the right unitary which is equivalent to the
  # U time signma (left unitary) * Singular values (Size of the PCs)
  pca_res <- prcomp(t(vst_top))


  # Plotting first two PCAs
  pca_plot <- plot_pca(bulk_dataset, PC_x = 1, PC_y = 2, color_by = color_variable)
```

# Pre-Clustering Check Point Question

**Checkpoint question:** Do any clinical labels show visible separation in reduced-dimensional space before clustering? What does this suggest about the transcriptomic signal for that variable?

No clincal labels (age_group, PSA_group, gleason_score) or a pathological labels (ajcc_pathologic_t, ajcc_pathologic_n ) show distinct concordance with either the unstanded or the TPM data for TCGA-PRAD only data. The only variables that show close agreement was the miRNA variables: paper_mRNA_cluster and the RNA_Cluster variable: Those that are directly relateted to transcriptomic data.Normal data was not included. This pattern holds for both variance stabilized raw counts and Log Normalized TPM Data.

It suggests that the those variables that show little to now ordered clusterting relative to the transcriptomic signal do not coincide with possible transcriptomic subtypes.

Global PCA does not capture the clinical variance of TCGA-PRAD data.

### **Converting Summarize Object into AnnData –\> Python Results**

Switch Over to the jupyterlab Notebook for k-means analysis

Tested with High Sil score on scaled HVGs of log(TPM + 1) GEX data

Results Summary

Result: Found that centroids are essentially stacked on top. Even if those outliers were deleted it would not change the terrible clustering.

Overall we see weak intrinsic clustering structure of HVG GEX space alone.

Looking below this would extent further to clinical and pathological variables as well

sil scores are all at or below 0

Which shows that any potential clustering method within HVG GEX space would be worse if based on clinical or pathological vars using Kmeans.

However, That does not mean that TPM GEX is non-clusterable: Kmeans assumes spherical clusters \# K-Means uses simple straight-line distance (Euclidean distance). #This makes it work best when clusters are roughly circular (in 2D) or spherical (in higher dimensions) and have similar sizes. #It struggles with elongated, curved, or unevenly sized clusters. #Since out data is naturally continuous, Kmeans is a bad fit because we are forcing it in bordered clusters.

**END OF DIM REDUCTION NOTEBOOK**
