---
title: "Preprocessing"
output: html_document
---

# Read in Library

```{r,include =FALSE}
library(tidyverse)
library(devtools)
library(vroom)
library(GEOquery)
library(biomaRt)
library(dplyr)
library(org.Hs.eg.db)
library(DESeq2)
library(clusterProfiler)
library(RNAseqQC)
#library(bulkanalysis) # fix this later
#source("Project_1_function_lab.R") # fix this later
```

# Querying for RNA-Seq Dataset

### Acquiring Chosen IDs and Filtering for Covariates

Checking meta data for key features

Gleason Score

**Pathologic T Stage:** grades how far the cancer has anatomically spreads within a or beyond the prostate.Pathalogic T stage (pT2 vs. pT3/T4)

PSA level at diagnosis (Common threshold: \<10, 10–20, \>20 ng/mL)

Biochemical Recurrance status (Available as a binary label)

Source of Pathology/Molecular Report: <https://pmc.ncbi.nlm.nih.gov/articles/PMC4695400/#S14>

ajcc_pathologi_t

paper_PSA_preop:

Reviewed_Gleason_sum :

ETV1_status

ETV4_status

FOXA1_mut

Mutations

Fraction_genome_altered

AR_score

# Need justification for each of following Chosen Covariates.

Why would they be meaningful feature that sample embeddings cluster around?

```{r}
library(openxlsx)
pathology_path <- "./GDCdata/TCGA_PRAD_pathology.xlsx"
pathology_metadata <- read.xlsx(pathology_path)
pathology_metadata
```

```{r}
#pathology covariates
path_cov <- c("SAMPLE_ID", 
              "PATIENT_ID", 
              "Age", 
              "Race", 
              "Subtype",
              "Reviewed_Gleason_sum", 
              "ETV1_status", 
              "ETV4_status", 
              "FOXA1_mut", 
              "Mutations", 
              "Fraction_genome_altered", 
              "AR_score",
              "PSA_preop")

covariates <- c("ajcc_pathologic_t", "gleason_score", "paper_PSA_preop", "paper_AR_score","paper_FOXA1_mut", "paper_ETV1_status", "paper_ETV4_status")

updated_covariates <- c("ajcc_pathologic_t", 
                        "gleason_score", 
                        "paper_PSA_preop", 
                        "paper_AR_score",
                        "paper_FOXA1_mut", 
                        "paper_ETV1_status", 
                        "paper_ETV4_status",
                        "Mutations",
                        "Fraction_genome_altered")
```

```{r}
# ==========================
# Clinical covariates
# ==========================

clinical_covariates <- c(
  "age_at_diagnosis",
  "cases",
  "race",
  "gender",
  "paper_PSA_preop"
)

# ==========================
# Pathological covariates
# ==========================

pathological_covariates <- c(
  "gleason_score",
  "ajcc_pathologic_t",
  "ajcc_pathologic_n",
  "primary_diagnosis",
  "paper_Residual_tumor",
  "Subtype"
)

# ==========================
# Molecular covariates
# ==========================

molecular_covariates <- c(
  "paper_iCluster",
  "paper_mRNA_cluster",
  "paper_methylation_cluster",

  "paper_ERG_status",
  "paper_ETV1_status",
  "paper_ETV4_status",
  "paper_SPINK1_high",

  "paper_FOXA1_mut",
  "paper_SPOP_mut",
  "paper_TP53_mut",
  "paper_PTEN_mut",

  "paper_Mutations",
  "paper_Fraction_genome_altered",

  "paper_ABSOLUTE_purity",
  "paper_AR_score"
)

# ==========================
# Survival variables
# ==========================

survival_covariates <- c(
  "vital_status",
  "days_to_last_follow_up",
  "cause_of_death"
)

# ==========================
# Identifiers
# ==========================

id_covariates <- c(
  "PATIENT_ID",
  "SAMPLE_ID"
)

# ==========================
# Complete covariate list
# ==========================

all_covariates <- c(
  id_covariates,
  clinical_covariates,
  pathological_covariates,
  molecular_covariates,
  survival_covariates
)
```

Pulling query results for the TCGA-PRAD datasets

```{r}
library(TCGAbiolinks)

#grab the results query
#I need the actual data
query <- GDCquery(
  project = "TCGA-PRAD",
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
)
# Primary Tumor: Direct tumor samples
#Solid Tissue Normal: These patient samples are called "solid tissue normals" and are taken from normal tissues near the tumor
results <- getResults(query)
results
# FIX: extract submitter ID properly
colnames(results)[colnames(results) == "cases.submitter_id"] <- "PATIENT_ID"

# ensure pathology metadata is filtered correctly
pathology_metadata <- pathology_metadata %>%
  dplyr::select(any_of(path_cov))

pathology_metadata
# join
merged_meta <- results %>%
  left_join(pathology_metadata, by = "PATIENT_ID")

merged_meta
```

```{r}
View(results)
```

```{r}
#Using 
clean_df <- merged_meta[complete.cases(merged_meta [, c("PATIENT_ID", "Reviewed_Gleason_sum", "PSA_preop")]), ]


# Ending Df samples into 15
#still results of the results query and not the meta_data the SE object
clean_df_sample <- clean_df
clean_df_sample

#Download the 15 samples instead of packing in 554 different DataSets
barcodes <- clean_df_sample$cases

query_uuid <- GDCquery(
        project = "TCGA-PRAD",
        data.category = "Transcriptome Profiling",
        data.type = "Gene Expression Quantification",
        barcode = barcodes)
  
  
GDCdownload(query_uuid)
  
se <- GDCprepare(query_uuid)

#pulling the metadata from the downloaded tcga datasets
tcga_coldata <- as.data.frame(colData(se))
tcga_coldata


#rename the tcga_coldat colums to mere with pathology meta_data
colnames(tcga_coldata)[colnames(tcga_coldata) == "barcode"] <- "cases"

merged_coldata <- left_join(clean_df_sample, tcga_coldata, by = "cases")
```

```{r}
View(tcga_coldata)
```

```{r}
library(S4Vectors)
merged_coldata <- merged_coldata %>% 
              dplyr::select(any_of(all_covariates)) %>% 
              tibble::column_to_rownames("cases")

colData(se) <- DataFrame(merged_coldata)
```

I need to decide what are the essentiall covariates that the sample embeddings of these covariates cluster for.

Right now we have15 datasets, w/ 60600 genes (all types with variant distinctions) and 30 covariates reduced from 150 when downloading the summarized experiment object from the TCGNA PRAD project.

I need to figure out how the sample embeddings of these cluster together. Some clever dimensionality reduction is necessary.

Need to figure out the definitions of the data type for each matrix available in the Summarized Experiment data object.

Normalization in RNA-Seq raw counts: "adjusting gene expression data to account for technical variabilities introduced during sample preparation such as in RNA quantity, quality, and amplfilcation efficiency. Normalization is important to ensure that gene expression data accuralty reflect bioogical differences between samples, rather than technical artifacts.

**What is the data?**

**Unstranded:** Raw counts ignoring strand information

**Stranded_first:** Counts assuming a first-strand (reverse stranded, "-") library

**stranded_second:** counts assuming a second_strand (forward-stranded, "+") library

<https://blog.genewiz.com/stranded-vs-non-stranded-rna-seq>

**fpkm_unstranded:** (Fragments Per Kilobase Million) from unstranded counts

**tpm_unstranded:** (Transcripts Per Million = reads per kilobase per million (RPKM) / fpkm): percentage of RPKM/FPKM calc.

1.  Gene length is normalized first, followed by the normalization of the sequencin depth. Different order of normalizaiton cna have significant impact on results. When using TPM, the sum of the TPMs

fpkm_uq_unstranded: upper quartile normalized FPKRM

<https://www.rna-seqblog.com/rpkm-fpkm-and-tpm-clearly-explained/>

# Meta-Data Alterations

#Okay learned something new. DDS objects don't accept TPM. \# This is because it's going off of library size expression and not direct relative expression

```{r}

#need to clean up the data that we have
alt_coldata <- as.data.frame(colData(se)) %>%

  mutate(across(where(~ is.character(.x) | is.factor(.x)),
                ~ gsub(",", ".", as.character(.x)))) %>%

  mutate(age_at_diagnosis = as.numeric(age_at_diagnosis),
         paper_PSA_preop = as.numeric(paper_PSA_preop),
         paper_ABSOLUTE_purity = as.numeric(paper_ABSOLUTE_purity),
         paper_Fraction_genome_altered = as.numeric(paper_Fraction_genome_altered)
         ) %>%
  
  mutate(
    paper_AR_score = gsub("%", "", paper_AR_score),
    paper_AR_score = trimws(paper_AR_score),
    paper_AR_score = as.numeric(paper_AR_score)
    ) %>%
  
  mutate(age_at_diagnosis =
           ifelse(age_at_diagnosis > 100,
                  floor(age_at_diagnosis / 365),
                  age_at_diagnosis)) %>%

  mutate(across(where(is.character),
                ~ replace(.x, is.na(.x), "none")))



alt_coldata
```

### Altering for Continous and Discrete Variables

```{r}
#Group Continuous Variables
alt_coldata$age_group <- cut(
  alt_coldata$age_at_diagnosis,
  breaks = c(0, 50, 60, 70, 80, Inf),
  labels = c("<50", "50-60", "60-70", "70-80", "80+"),
  include.lowest = TRUE
)

alt_coldata$PSA_group <- cut(
  alt_coldata$paper_PSA_preop,
  breaks = c(0, 5, 10, 15, 20, Inf),
  labels = c("<5", "5-10", "10-15", "15-20", "20+"),
  include.lowest = TRUE
)

alt_coldata$AR_score_group <- cut(
  alt_coldata$paper_AR_score,
  breaks = c(-20, -10 ,-5 ,0, 25, 50, 75, 100, Inf),
  labels = c("-20 - -10","-10 - -5","-5 - 0"," 0-25", "25-50", "50-75", "75-100", "100+"),
  include.lowest = TRUE
)

alt_coldata$ABSOLUTE_purity_group <- cut(
  alt_coldata$paper_ABSOLUTE_purity,
  breaks = c(0, 0.50, 0.70, 0.85, 1.00),
  include.lowest = TRUE,
  labels = c("Low", "Moderate", "High", "Very High")
)

alt_coldata$FGA_group <- cut(
  alt_coldata$paper_Fraction_genome_altered,
  breaks = c(0, 0.25, 0.50, 0.75, 1.00),
  include.lowest = TRUE,
  labels = c("0-0.25", "0.25-0.50", "0.50-0.75", "0.75-1.00")
)



alt_coldata
colData(se) <- DataFrame(alt_coldata)
```

```{r}
View(alt_coldata)
```

TCGA-PRAD: Project dataformat

m x n matrix where is a column wise samples. m = 60k genes, and n = \# of barcodes/patient/samples requested from the database.

Essentially will be classfiying on the column basis instead of the dataset basis makes sense. I am used to seeing it as an mxn problem not an columun wise column.

Goal is classify each of these into a clinical label.

**Input:** TCGA Transcriptomic data Column, and the x amount of variable included for each of those columns of count data

**Output:** A clinical Label classification based on those effects

**Full Project Question:** Do molecular subtypes defined from transcriptomic data align with clincal classification?

How does molecular sub-typing of PRAD relate to the clinical classification ? How does transcriptomic data it help more or better than the clinical variables that are collected?

What makes PRAD soooo clinically heterogenous: "patients with similar histological grade can have vastly different outcomes"

How does molecular subtyping of PRAD relate to the clinical classification or at least inform clinical classification?

My job:

1.  Perform exploratory dim reduction an clustering for transcriptomic subtypes
2.  Use those as labels for the data per column/patient/sample
3.  Build the classifier based on the transcriptomic subtypes

## AnnData Conversion

```{r}
library(convert2anndata)
library(Seurat)
```

```{r}
#transfer the rownames of the sample onto the unstranded matrix
for (i in seq_along(assays(se))) {
  
  mat <- assays(se, withDimnames=FALSE)[[i]]
  
  colnames(mat) <- rownames(colData(se))
  rownames(mat) <- rownames(as.data.frame(rowRanges(se)))

  assays(se)[[i]] <- mat
  
}

#Okay learned something new. DDS objects don't accept TPM and Seurat also hates TPM normalized counts.
# This is because it's going off of library size expression and not direct relative expression
# Intersting Article about Seurat basics: https://github.com/satijalab/seurat/issues/7496
```

### Converting SE into h5ad

```{r}
library(convert2anndata)
library(anndata)
```

```{r, include = FALSE}
ad_path <- "./GDCdata/"

seurat_se <-  CreateSeuratObject(
  counts = assays(se)[["unstranded"]],
  meta.data = colData(se)
  
)
#Log noralizing unstranded data
seurat_se <- NormalizeData(seurat_se)
seurat_se <- FindVariableFeatures(seurat_se, nfeatures = 5000)
seurat_se <- ScaleData(seurat_se)
seurat_se <- RunPCA(seurat_se)
sce <- convert_seurat_to_sce(seurat_se)
ad <- convert_to_anndata(sce, assayName = "counts", useAltExp = TRUE)
write_h5ad(ad, paste0(ad_path, "seurat_se.h5ad"))


#Unfortunetly I need to create a seperate Seurat Assay
counts <- assays(se)[["tpm_unstrand"]]
seurat_se_tpm <- CreateSeuratObject(
  counts = counts,
  meta.data = as.data.frame(colData(se))
)
seurat_se_tpm <- NormalizeData(seurat_se_tpm, normalization.method = "LogNormalize")
seurat_se_tpm <- FindVariableFeatures(seurat_se_tpm, nfeatures = 5000)
seurat_se_tpm <- ScaleData(seurat_se_tpm)
seurat_se_tpm <- RunPCA(seurat_se_tpm)

sce_tpm  <- convert_seurat_to_sce(seurat_se_tpm)

ad_tpm <- convert_to_anndata(sce_tpm, assayName = "counts", useAltExp = TRUE)

write_h5ad(ad_tpm, paste0(ad_path, "seurat_se_tpm.h5ad"))
```

```{r}
#saving Curated Summarized Experiment
save_data <- "saved_data"
if(!dir.exists(save_data)) dir.create(save_data, recursive = TRUE)
saveRDS(se, "./saved_data/curated_RSE_object.rds")
```

**END OF PRE-PROCESSING NOTEBOOK**

# **DimReduction TPM + Log2 Norm + Metadata coloring**

Reading in curated data file

```{r}
se <- readRDS("./saved_data/curated_RSE_object.rds")
```

```{r}
met <- as.data.frame(colData(se))
rownames <- rownames(as.data.frame(colData(se)))
met$SAMPLE_ID <- rownames
rownames(met) <- rownames(NULL)

met
```

```{r}
tpm.mat <- se@assays@data$tpm_unstrand

log_tpm <- log2(tpm.mat + 1)


#Select top variable gene
top_var_genes <- head(order(rowVars(log_tpm), decreasing = TRUE), 5000)
vst_top <- log_tpm[top_var_genes, ]
# Run PCA via R
# Remember PCA is taking the right unitary martrix (VT) (feature space) of SVD
# VT tells us the principal directions of the data set
# multiply the the original matrix by the right unitary which is equivalent to the
# U time signma (left unitary) * Singular values (Size of the PCs)


#PCA on genes
pca_res <- prcomp(
  t(log_tpm[top_var_genes, ]),
  center = TRUE,
  scale. = TRUE)

pca_df <- data.frame(
  SAMPLE_ID = colnames(vst_top),
  PC1= pca_res$x[ ,1],
  PC2 = pca_res$x[ ,2]
)
#got a duplicate error. Don't know why R decided it's a problem now but had to make a workaround
met <- as.data.frame(colData(se))
rownames <- rownames(as.data.frame(colData(se)))
met$SAMPLE_ID <- rownames
rownames(met) <- rownames(NULL)
pca_df <- left_join(pca_df, met , by = "SAMPLE_ID")
```

```{r}
vars <- colnames(pca_df)[5:ncol(pca_df)]
PCA_plots <- list()
for (v in vars) {

  p <- ggplot(
    pca_df,
    aes(PC1, PC2, color = factor(.data[[v]]))
  ) +
    geom_point() +
    ggtitle(glue::glue("PCA space with Var: {v}"))

  PCA_plots[[v]] <- p
}


 # Calculating Percentage of Variance Per PC for the Scree Plot
  
  # Calculate percentage variance explained
  var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100
  
  # Create a data frame for plotting (first 10 PCs)
  scree_data <- data.frame(
    PC = paste0("PC", 1:10),
    Variance = var_explained[1:10]
  )
  scree_data
  # Ensure PC is treated as a factor in order
  scree_data$PC <- factor(scree_data$PC, levels = scree_data$PC)
  
  # Scree Plot ensures that we are plotting PCs that capture the most amount of variability within our data
  scree_plot <- ggplot(scree_data, aes(x = PC, y = Variance, group = 1)) +
    geom_line(color = "red", size = 1) +
    geom_point(color = "red", size = 2) +
    labs(title = "Scree Plot: Variance Explained per PC",
         x = "Principal Component",
         y = "Percentage of Variance (%)") +
    theme_minimal()

scree_plot
```

```{r}

#Clinical Covariates
PCA_plots[["PSA_group"]]
PCA_plots[["gleason_score"]]
PCA_plots[["ajcc_pathologic_t"]]
PCA_plots[["ajcc_pathologic_n"]]
PCA_plots$Subtype

#molecular covariates
PCA_plots[["paper_ETV1_status"]]
PCA_plots[["paper_FOXA1_mut"]]
PCA_plots[["FGA_group"]]
PCA_plots$paper_mRNA_cluster
PCA_plots$paper_iCluster
PCA_plots$AR_score_group
PCA_plots[["FGA_group"]]
PCA_plots$paper_ETV1_status
PCA_plots$paper_TP53_mut
```

# Unstranded Counts + VST + PCA

```{r, include = FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
ah[["AH89426"]]
```

```{r, include = FALSE}
# Unflitered Counts Saved to a DeSeqDataSet (dds object)
tcgna_prad <- make_dds(counts = se@assays@data$unstranded , metadata = colData(se), ah_record = "AH89426")
#renaming counts to raw counts
assayNames(tcgna_prad)[1] <- "counts"

#variance stabilized Raw counts
assays(tcgna_prad)[[2]] <- vst(tcgna_prad@assays@data$counts)
assayNames(tcgna_prad)[2] <- "var_stable"
#log_transformed (log normalized counts)
assays(tcgna_prad)[[3]] <- log10(tcgna_prad@assays@data$counts + 1)
assayNames(tcgna_prad)[3] <- "log_counts"
```

```{r}
bulk_dataset <- tcgna_prad
anno_variable <- "gleason_score"
color_variable <-  "gleason_score"
#prioritize variance stable assay within the bulk_dataset
  assays(bulk_dataset) <- assays(bulk_dataset)[c("var_stable", "counts", "log_counts")]
  # takes the euclidan distance of a variance stabilized dataset
  # row by row euclidean distance against each other
  # plots the visual distance between each sample onto heatmap
  distance_plot <- plot_sample_clustering(bulk_dataset, anno_vars = anno_variable, distance = "euclidean")


  #____Extracting Top Variable Gene per sample -----#
  # Setting Variance Stable Counts as Priority Assay
  vst_mat <- assay(bulk_dataset, "var_stable")
  #Select top variable gene
  top_var_genes <- head(order(rowVars(vst_mat), decreasing = TRUE), 2000)
  vst_top <- vst_mat[top_var_genes, ]
  # Run PCA via R
  # Remember PCA is taking the right unitary martrix (VT) (feature space) of SVD
  # VT tells us the principal directions of the data set
  # multiply the the original matrix by the right unitary which is equivalent to the
  # U time signma (left unitary) * Singular values (Size of the PCs)
  pca_res <- prcomp(t(vst_top))


  # Plotting first two PCAs
  pca_plot <- plot_pca(bulk_dataset, PC_x = 1, PC_y = 2, color_by = color_variable)
```

## Pre-Clustering Check Point Question

**Checkpoint question:** Do any clinical labels show visible separation in reduced-dimensional space before clustering? What does this suggest about the transcriptomic signal for that variable?

No clincal labels (age_group, PSA_group, gleason_score) or a pathological labels (ajcc_pathologic_t, ajcc_pathologic_n ) show distinct concordance with either the unstanded or the TPM data for TCGA-PRAD only data. The only variables that show close agreement was the miRNA variables: paper_mRNA_cluster and the RNA_Cluster variable: Those that are directly relateted to transcriptomic data.Normal data was not included. This pattern holds for both variance stabilized raw counts and Log Normalized TPM Data.

It suggests that the those variables that show little to now ordered clusterting relative to the transcriptomic signal do not coincide with possible transcriptomic subtypes.

Global PCA does not capture the clinical variance of TCGA-PRAD data.

### **Converting Summarize Object into AnnData –\> Python Results**

Tested with High Sil score on scaled HVGs of log(TPM + 1) GEX data

Results Summary

Result: Found that centroids are essentially stacked on top. Even if those outliers were deleted it would not change the terrible clustering.

Overall we see weak intrinsic clustering structure of HVG GEX space alone.

Looking below this would extent further to clinical and pathological variables as well

sil scores are all at or below 0

Which shows that any potential clustering method within HVG GEX space would be worse if based on clinical or pathological vars using Kmeans.

However, That does not mean that TPM GEX is non-clusterable: Kmeans assumes spherical clusters \# K-Means uses simple straight-line distance (Euclidean distance). #This makes it work best when clusters are roughly circular (in 2D) or spherical (in higher dimensions) and have similar sizes. #It struggles with elongated, curved, or unevenly sized clusters. #Since out data is naturally continuous, Kmeans is a bad fit because we are forcing it in bordered clusters.

# HClust on log(TPM + 1) HVG PCA space

In another python notebook we looked into k-means clustering based on log(TPM + 1) based on solely transcriptome data

Problems:

1\. TCGA-PRAD is continous

2\. K-Means uses simple straight-line distance (Euclidean distance).

3\. This makes it work best when clusters are roughly circular (in 2D) or spherical (in higher dimensions) and have similar sizes.

4\. It struggles with elongated, curved, or unevenly sized clusters.

5\. Since out data is naturally continuous, Kmeans is a bad fit because we are forcing it in bordered clusters.

```{r}
# Define clinical covariates
clinical_covariates <- c(
  "PSA_group",
  "gleason_score",
  "ajcc_pathologic_t",
  "ajcc_pathologic_n",
  "Subtype"
)

# Define molecular covariates (duplicates removed)
molecular_covariates <- c(
  "paper_ETV1_status",
  "paper_FOXA1_mut",
  "FGA_group",
  "paper_mRNA_cluster",
  "paper_iCluster",
  "AR_score_group",
  "paper_TP53_mut"
)

hclusters <- c("hclust")
# Combine into a single vector if needed for your loop
relevant_covs <- c(clinical_covariates, molecular_covariates, hclusters)
```

```{r}
library(irlba)
# keep irlba in this your backpocket just in case this doesn't work out (Truncated SVD) --> Does Using the Left Unitary matrix along help us ? I doubt because we have continous data
```

```{r}
top_pcs <- pca_res$x[, 1:10]

dist_obj <- dist(top_pcs, method = "euclidean")
hc_out <- hclust(dist_obj, method = "ward.D2")
# within total sum of squares
max_k <- 20

cluster_eval <- do.call(
  rbind,
  lapply(1:max_k, function(k) {
    km <- kmeans(as.matrix(top_pcs), centers = k, nstart = 25)
    
    data.frame(
      k = k,
      total_wss = km$tot.withinss,
      wss_tss = km$tot.withinss / km$totss)
  })
)
  
plot(1:max_k, cluster_eval$total_wss, type="b", pch = 19, 
  xlab="Number of clusters K",
   ylab="Total WSS",
   main="Total WSS vs K Clusters")

plot(1:max_k, cluster_eval$wss_tss, type="b", pch = 19, 
   xlab="Number of clusters K",
   ylab="Total within-clusters / total  sum of squares",
   main="Total WSS/TSS vs K Clusters")

clusters <- cutree(hc_out, k = 10)

#CHECK-IN TIME:

#before plotting this let's analyze what we are doing and what we expect to happen
#What we have: pca coordinates of log(TPM + 1) of top 5000 HVGs dim reduced by PCA Space

#What did we do to the data: Used Total Within Sum of Squares and WSS/TSS to estimate how many clusters we can use after a hierarcical clustering performance

# Result: smooth, non-elbow curve

# We then used hclust on those PCA coordinates (U * Sigma) scaled sample embeddings

# Why are we using Hclust ?: Because that is usually what is used for bulk-RNAseq data
# K-means Clustering was already tried and could not resolve continuous data into functional discrete clusters

# What do we expect (visually): long chain/ weak branching, not likely to yield discrete clusters for any optimized K value
# https://projector-video-pdf-converter.datacamp.com/18542/chapter3.pdf

pca_df_hclust <- data.frame(
  SAMPLE_ID = colnames(vst_top),
  PC1= pca_res$x[ ,1],
  PC2 = pca_res$x[ ,2],
  hclust = clusters
)

pca_df_hclust <- left_join(pca_df_hclust, met , by = "SAMPLE_ID")
#before plotting this
pca_df_hclust

vars <- colnames(pca_df_hclust)[colnames(pca_df_hclust) %in% relevant_covs]
PCA_plots <- list()
for (v in vars) {
  
  p <- ggplot(
    pca_df_hclust,
    aes(PC1, PC2, color = factor(.data[[v]]))
  ) +
    geom_point() +
    ggtitle(glue::glue("PCA space with Var: {v}"))
  
  PCA_plots[[v]] <- p
}
PCA_plots$hclust
```

#CHECK-IN TIME:

#before plotting this let's analyze what we are doing and what we expect to happen #What we have: pca coordinates of log(TPM + 1) of top 5000 HVGs dim reduced by PCA Space

#What did we do to the data: Used Total Within Sum of Squares and WSS/TSS to estimate how many clusters we can use after a hierarcical clustering performance

**Result:** smooth, non-elbow curve for both WSS and WSS/TSS

We then used hclust on those PCA coordinates (U \* Sigma) scaled sample embeddings

**Why are we using Hclust ?**: Because that is usually what is used for bulk-RNAseq data

K-means Clustering was already tried and could not resolve continuous data into functional discrete clusters

What do we expect (visually): long chain/ weak branching, not likely to yield discrete clusters for any optimized K value

Result: arbituarty stratification of clusters after Hclust within PC Space. Used WSS and WSS/TSS to asses possible clusters. Found smooth decreases

We clearly have continuous latent factors that are much greater than discrete clusters.

```{r}
meta_data_cols <- colnames(pca_df_hclust)[9:ncol(pca_df_hclust)]

top_pcs <- pca_res$x[, 1:10]

meta_data <-pca_df_hclust %>% select(any_of(meta_data_cols))

meta_data[] <- lapply(meta_data, factor)


get_p_value <- function(pc_vector, var) {
  
  
  valid <- !is.na(pc_vector) & !is.na(var)
  
  pc_vector <- pc_vector[valid]
  var_vector <- var[valid]
  
  
  if (is.numeric(var_vector)) {
    # Continuous variable: use linear regression/correlation test
    return(summary(lm(pc_vector ~ var_vector))$coefficients[2, 4])
    
  } else {
    
    # Categorical variable: use ANOVA
    fit <- aov(pc_vector ~ as.factor(var_vector))
    
    anova_sum <- summary(fit)[[1]]
    p_val <- anova_sum[1, "Pr(>F)"]
    
    if (length(p_val) != 1 || is.na(p_val) || is.nan(p_val) || is.infinite(p_val)) {
      return(NA)
    }
    
    
    return(p_val)
  }
}


p_matrix <- matrix(NA, 
                   nrow = length(meta_data), 
                   ncol = ncol(top_pcs),
                   dimnames = list(names(meta_data), colnames(top_pcs)))



```

```{r}
library(glue)
for (v in names(meta_data)) {
  
  current_var <- meta_data[[v]]
  
  # compute number of levels safely
  n_levels <- if (is.factor(current_var) || is.character(current_var)){
    nlevels(as.factor(current_var))
  } else {
    length(unique(na.omit(current_var)))
  }
  # skip uninformative variables
  if (n_levels < 2) next
  
  for (pc in colnames(top_pcs)) {
    
    current_pc <- top_pcs[, pc]
    
    # safe p-value calculation (assumes your function handles NA)
    p_matrix[v, pc] <- get_p_value(current_pc, current_var)
  }
}

# apply logarithmic scaling to pval matrix
log_p_matrix <- -log10(p_matrix)

# account for inifinite ones
max_finite <- max(log_p_matrix[is.finite(log_p_matrix)])

# use if infinite + 1 to. force discrete variable
log_p_matrix[is.infinite(log_p_matrix)] <- max_finite + 1

# apply a discrete force to possible na values
log_p_matrix <- log_p_matrix[!apply(is.na(p_matrix), 1 , all), ]

# arrange variables
# Order rows by strongest association across PCs
row_score <- apply(log_p_matrix, 1, max, na.rm = TRUE)

log_p_matrix <- log_p_matrix[
  order(row_score, decreasing = TRUE),
]
# 5. Plot the heatmap
p_val_map <- pheatmap::pheatmap(log_p_matrix, 
                                legend_labels = "log10(pvalue)",
                                cluster_rows=F, 
                                cluster_cols=F, 
                                main = "PC-Metadata Associations (-log10(p-value))",
                                )
```

Sanity Check Time

**Initial Question:** Do molecular subtypes defined from transcriptomic data align with established clinical classifications?

**Task:** You will download TCGA prostate adenocarcinoma (PRAD) data, perform unsupervised clustering to discover transcriptomic subtypes, evaluate how well those subtypes correspond to clinical labels, and build a supervised classifier.

**Reccomended Workflow:**

Download Data –\> Curation -\> normalize to CPM, TPM, or FPKM

Dim Reduce with PCA and UMAP, visualise groupings of clinical,pathological, and molecular variables –\> Visualize

Unsupervised clustering

**Recapping Workflow.**

1.  Data Curation –\> 232 TCGA-PRAD samples x 60600 genes

2.  Meta_data_cleaning -\> Binned continuous variables (age_group, PSA score, AR score, ABSOLUTE purity, Fraction_genome_altered)

3.   Dim Reduction –\> log(TPM of HVGs(5k) +1 ) –\> PCA (centered and scaled Data) + Scree Plot for Variance Ratio per top PC –\> colored embeddings by 30 covariated (clinical, pathological, molecular, genomic)

4.   Unsupervised Clustering part 1: Kmeans (sklearn in python): estimated average silhouette score for clusterting based on log(TPM of HVGs(5k)+1 ) and performed kmeans clustering –\> result: Overall we see weak intrinsic clustering structure of HVG GEX space alone.

    Performed Sil score for kmeans clustering on metadata: found scores ranging from -0.072 to 0.009

5.   Unsupervised Cluster part 2:

    Input: Top 10 PCs of log(TPM of HVGs(5k) + 1) GEX data

    Hclustering on PCs: Graphed WSS/TSS and totalWSS to estimate cluster cut off after Hclust –\> performed Hclust with Euclidean Distance (L2 Norm)

    Graphed WSS/TSS and totalWSS to estimate cluster cut off for Hclust –\> found a smooth elbow plot for both determining which means that there is not much cluster separation or tightness

    Its clear that there are continous latent variables within our disease homongenous data. Unlinke my previous bulk-RNAseq data, there are not two disease states only one.

6.   Moved to PVCA for variance association to find the variables that are associated with variance within the Top 10 PCs of log(TPM of HVGs(5k) + 1) GEX data:

    Result:

    **Highest Variance Associations Per PC**

    PC1: paper_iCluster, paper_mRNA_cluster, Absolute_purity_group, paper_Residual tumor

    PC2: at best Subtype, paper_Residual_tumor

    PC4: paper_methylation_cluster, paper_mRNA_cluster, AR_score_group

    PC5: mRNA_cluster, paper_methylation_cluster

    PC6: paper_iCluster,paper_methylation_cluster, paper_ERG_Status

    Biggest Note: None of clinical variables have direct association with the variance within the PC space (paper_PSA/PSA_group, ajcc_pathologict_t, gleason score)

    Subtype is not associated with the Variance in PC1 but PC6. Right before the drop off to PC7-PC10 where the scree plot show Percentage of variance of the data flatten to 0.

**What we know:**

We know that the clinical and pathological variables very much are mis-aligned with the Transcriptomic signals even with scaling and normalization within the PC space.

We know which variables do in fact align with the data somewhat: paper_iCluster, paper_mRNA_cluster, Abosulte_cluster, AR_score_group, paper_methylation_cluster, paper_ERG Status

What does this mean? and Have papers Previsouly noted this ?

The clinical classifcations of PRAD like Gleason Score, PSA levels, and AJCC_T-stage are human-defined categories based on STRUCTURAL tissue architechture and clinical risk. They do not map cleanly to global mRNA expression cascades noted in this beutifal paper <https://pubmed.ncbi.nlm.nih.gov/26544944/>

Huge task; Note down what each Variable Means. If writing report this should go at the very top of the paper.

**Initial Questions:**

Now what?

PVCA related question: why does binned continuos variables capture more variation within the GEX data compared to raw variables?

Should I have completeley ommited the NAs from this data: IMO no this makes the data more realistic?

Would it be wise for Combat-Seq batch-correction? would this condense meta-variables to PC1 ? How would this change variance ratio of the PCS?

There is a paper that uses ConsenusClustering: <https://pubmed.ncbi.nlm.nih.gov/33903734/> (Likely next step to verify continuous latent variables) –\> Recc Reading: Han et al. 2021, and Wilkerson, Hayes (2021): Consensus Clustering implementation and how to interpret consensus matrices, track items and use CDF plots. for cluster stabiity

How should this data be labeled for the classifier ? If we do build one The focus will obviously be heavy on dim reduction utilizing the left unitary matrix

It's likely that we will overfit hard on this Data we'll likely need a mixture of probabilistic and LR + random Forest this. The decision boundary is going to be a pain the ass. I know somehow I'll end up using a deep-learning method.

I don't really have an end goal for this? Would it be to build a supervised classifier for prostate cancer data?

Demonstrate how anchoring a classifier to latent continuous variables allows ML to extract independent, clincally predictive GEX signatures

Possible Moves for the classifier: SVM with Radial Basis Function/ Gausian Kernel
