---
title: "Project_2_UnsupervisedClust_part2"
output: html_document
---

```{r,include =FALSE}
library(tidyverse)
library(devtools)
library(vroom)
library(GEOquery)
library(biomaRt)
library(dplyr)
library(org.Hs.eg.db)
library(DESeq2)
library(clusterProfiler)
library(RNAseqQC)
#library(bulkanalysis) # fix this later
#source("Project_1_function_lab.R") # fix this later
```

Read in Curated Data Object

```{r}
se <- readRDS("./saved_data/curated_RSE_object.rds")
```

# HClust on log(TPM + 1) HVG PCA space

In another python notebook we looked into k-means clustering based on log(TPM + 1) based on solely transcriptome data

Problems:

1\. TCGA-PRAD is continous

2\. K-Means uses simple straight-line distance (Euclidean distance).

3\. This makes it work best when clusters are roughly circular (in 2D) or spherical (in higher dimensions) and have similar sizes.

4\. It struggles with elongated, curved, or unevenly sized clusters.

5\. Since out data is naturally continuous, Kmeans is a bad fit because we are forcing it in bordered clusters.

```{r}
# Define clinical covariates
clinical_covariates <- c(
  "PSA_group",
  "gleason_score",
  "ajcc_pathologic_t",
  "ajcc_pathologic_n",
  "Subtype"
)

# Define molecular covariates (duplicates removed)
molecular_covariates <- c(
  "paper_ETV1_status",
  "paper_FOXA1_mut",
  "FGA_group",
  "paper_mRNA_cluster",
  "paper_iCluster",
  "AR_score_group",
  "paper_TP53_mut"
)

hclusters <- c("hclust")
# Combine into a single vector if needed for your loop
relevant_covs <- c(clinical_covariates, molecular_covariates, hclusters)
```

```{r}
library(irlba)
# keep irlba in this your backpocket just in case this doesn't work out (Truncated SVD) --> Does Using the Left Unitary matrix along help us ? I doubt because we have continous data
```

```{r}
top_pcs <- pca_res$x[, 1:10]

dist_obj <- dist(top_pcs, method = "euclidean")
hc_out <- hclust(dist_obj, method = "ward.D2")
# within total sum of squares
max_k <- 20

cluster_eval <- do.call(
  rbind,
  lapply(1:max_k, function(k) {
    km <- kmeans(as.matrix(top_pcs), centers = k, nstart = 25)
    
    data.frame(
      k = k,
      total_wss = km$tot.withinss,
      wss_tss = km$tot.withinss / km$totss)
  })
)
  
plot(1:max_k, cluster_eval$total_wss, type="b", pch = 19, 
  xlab="Number of clusters K",
   ylab="Total WSS",
   main="Total WSS vs K Clusters")

plot(1:max_k, cluster_eval$wss_tss, type="b", pch = 19, 
   xlab="Number of clusters K",
   ylab="Total within-clusters / total  sum of squares",
   main="Total WSS/TSS vs K Clusters")

clusters <- cutree(hc_out, k = 10)

#CHECK-IN TIME:

#before plotting this let's analyze what we are doing and what we expect to happen
#What we have: pca coordinates of log(TPM + 1) of top 5000 HVGs dim reduced by PCA Space

#What did we do to the data: Used Total Within Sum of Squares and WSS/TSS to estimate how many clusters we can use after a hierarcical clustering performance

# Result: smooth, non-elbow curve

# We then used hclust on those PCA coordinates (U * Sigma) scaled sample embeddings

# Why are we using Hclust ?: Because that is usually what is used for bulk-RNAseq data
# K-means Clustering was already tried and could not resolve continuous data into functional discrete clusters

# What do we expect (visually): long chain/ weak branching, not likely to yield discrete clusters for any optimized K value
# https://projector-video-pdf-converter.datacamp.com/18542/chapter3.pdf

pca_df_hclust <- data.frame(
  SAMPLE_ID = colnames(vst_top),
  PC1= pca_res$x[ ,1],
  PC2 = pca_res$x[ ,2],
  hclust = clusters
)

pca_df_hclust <- left_join(pca_df_hclust, met , by = "SAMPLE_ID")
#before plotting this
pca_df_hclust

vars <- colnames(pca_df_hclust)[colnames(pca_df_hclust) %in% relevant_covs]
PCA_plots <- list()
for (v in vars) {
  
  p <- ggplot(
    pca_df_hclust,
    aes(PC1, PC2, color = factor(.data[[v]]))
  ) +
    geom_point() +
    ggtitle(glue::glue("PCA space with Var: {v}"))
  
  PCA_plots[[v]] <- p
}
PCA_plots$hclust
```

#CHECK-IN TIME:

#before plotting this let's analyze what we are doing and what we expect to happen #What we have: pca coordinates of log(TPM + 1) of top 5000 HVGs dim reduced by PCA Space

#What did we do to the data: Used Total Within Sum of Squares and WSS/TSS to estimate how many clusters we can use after a hierarcical clustering performance

**Result:** smooth, non-elbow curve for both WSS and WSS/TSS

We then used hclust on those PCA coordinates (U \* Sigma) scaled sample embeddings

**Why are we using Hclust ?**: Because that is usually what is used for bulk-RNAseq data

K-means Clustering was already tried and could not resolve continuous data into functional discrete clusters

What do we expect (visually): long chain/ weak branching, not likely to yield discrete clusters for any optimized K value

Result: arbituarty stratification of clusters after Hclust within PC Space. Used WSS and WSS/TSS to asses possible clusters. Found smooth decreases

We clearly have continuous latent factors that are much greater than discrete clusters.

```{r}
meta_data_cols <- colnames(pca_df_hclust)[9:ncol(pca_df_hclust)]

top_pcs <- pca_res$x[, 1:10]

meta_data <-pca_df_hclust %>% select(any_of(meta_data_cols))

meta_data[] <- lapply(meta_data, factor)


get_p_value <- function(pc_vector, var) {
  
  
  valid <- !is.na(pc_vector) & !is.na(var)
  
  pc_vector <- pc_vector[valid]
  var_vector <- var[valid]
  
  
  if (is.numeric(var_vector)) {
    # Continuous variable: use linear regression/correlation test
    return(summary(lm(pc_vector ~ var_vector))$coefficients[2, 4])
    
  } else {
    
    # Categorical variable: use ANOVA
    fit <- aov(pc_vector ~ as.factor(var_vector))
    
    anova_sum <- summary(fit)[[1]]
    p_val <- anova_sum[1, "Pr(>F)"]
    
    if (length(p_val) != 1 || is.na(p_val) || is.nan(p_val) || is.infinite(p_val)) {
      return(NA)
    }
    
    
    return(p_val)
  }
}


p_matrix <- matrix(NA, 
                   nrow = length(meta_data), 
                   ncol = ncol(top_pcs),
                   dimnames = list(names(meta_data), colnames(top_pcs)))



```

```{r}
library(glue)
for (v in names(meta_data)) {
  
  current_var <- meta_data[[v]]
  
  # compute number of levels safely
  n_levels <- if (is.factor(current_var) || is.character(current_var)){
    nlevels(as.factor(current_var))
  } else {
    length(unique(na.omit(current_var)))
  }
  # skip uninformative variables
  if (n_levels < 2) next
  
  for (pc in colnames(top_pcs)) {
    
    current_pc <- top_pcs[, pc]
    
    # safe p-value calculation (assumes your function handles NA)
    p_matrix[v, pc] <- get_p_value(current_pc, current_var)
  }
}

# apply logarithmic scaling to pval matrix
log_p_matrix <- -log10(p_matrix)

# account for inifinite ones
max_finite <- max(log_p_matrix[is.finite(log_p_matrix)])

# use if infinite + 1 to. force discrete variable
log_p_matrix[is.infinite(log_p_matrix)] <- max_finite + 1

# apply a discrete force to possible na values
log_p_matrix <- log_p_matrix[!apply(is.na(p_matrix), 1 , all), ]

# arrange variables
# Order rows by strongest association across PCs
row_score <- apply(log_p_matrix, 1, max, na.rm = TRUE)

log_p_matrix <- log_p_matrix[
  order(row_score, decreasing = TRUE),
]
# 5. Plot the heatmap
p_val_map <- pheatmap::pheatmap(log_p_matrix, 
                                legend_labels = "log10(pvalue)",
                                cluster_rows=F, 
                                cluster_cols=F, 
                                main = "PC-Metadata Associations (-log10(p-value))",
                                )
```

Sanity Check Time

**Initial Question:** Do molecular subtypes defined from transcriptomic data align with established clinical classifications?

**Task:** You will download TCGA prostate adenocarcinoma (PRAD) data, perform unsupervised clustering to discover transcriptomic subtypes, evaluate how well those subtypes correspond to clinical labels, and build a supervised classifier.

**Reccomended Workflow:**

Download Data –\> Curation -\> normalize to CPM, TPM, or FPKM

Dim Reduce with PCA and UMAP, visualise groupings of clinical,pathological, and molecular variables –\> Visualize

Unsupervised clustering

**Recapping Workflow.**

1.  Data Curation –\> 232 TCGA-PRAD samples x 60600 genes

2.  Meta_data_cleaning -\> Binned continuous variables (age_group, PSA score, AR score, ABSOLUTE purity, Fraction_genome_altered)

3.  Dim Reduction –\> log(TPM of HVGs(5k) +1 ) –\> PCA (centered and scaled Data) + Scree Plot for Variance Ratio per top PC –\> colored embeddings by 30 covariated (clinical, pathological, molecular, genomic)

4.  Unsupervised Clustering part 1: Kmeans (sklearn in python): estimated average silhouette score for clusterting based on log(TPM of HVGs(5k)+1 ) and performed kmeans clustering –\> result: Overall we see weak intrinsic clustering structure of HVG GEX space alone.

```         
Performed Sil score for kmeans clustering on metadata: found scores ranging from -0.072 to 0.009
```

5.  Unsupervised Cluster part 2:

```         
Input: Top 10 PCs of log(TPM of HVGs(5k) + 1) GEX data

Hclustering on PCs: Graphed WSS/TSS and totalWSS to estimate cluster cut off after Hclust –\> performed Hclust with Euclidean Distance (L2 Norm)

Graphed WSS/TSS and totalWSS to estimate cluster cut off for Hclust –\> found a smooth elbow plot for both determining which means that there is not much cluster separation or tightness

Its clear that there are continous latent variables within our disease homongenous data. Unlinke my previous bulk-RNAseq data, there are not two disease states only one.
```

6.  Moved to PVCA for variance association to find the variables that are associated with variance within the Top 10 PCs of log(TPM of HVGs(5k) + 1) GEX data:

```         
Result:

**Highest Variance Associations Per PC**

PC1: paper_iCluster, paper_mRNA_cluster, Absolute_purity_group, paper_Residual tumor

PC2: at best Subtype, paper_Residual_tumor

PC4: paper_methylation_cluster, paper_mRNA_cluster, AR_score_group

PC5: mRNA_cluster, paper_methylation_cluster

PC6: paper_iCluster,paper_methylation_cluster, paper_ERG_Status

Biggest Note: None of clinical variables have direct association with the variance within the PC space (paper_PSA/PSA_group, ajcc_pathologict_t, gleason score)

Subtype is not associated with the Variance in PC1 but PC6. Right before the drop off to PC7-PC10 where the scree plot show Percentage of variance of the data flatten to 0.
```

**What we know:**

We know that the clinical and pathological variables very much are mis-aligned with the Transcriptomic signals even with scaling and normalization within the PC space.

We know which variables do in fact align with the data somewhat: paper_iCluster, paper_mRNA_cluster, Abosulte_cluster, AR_score_group, paper_methylation_cluster, paper_ERG Status

What does this mean? and Have papers Previsouly noted this ?

The clinical classifcations of PRAD like Gleason Score, PSA levels, and AJCC_T-stage are human-defined categories based on STRUCTURAL tissue architechture and clinical risk. They do not map cleanly to global mRNA expression cascades noted in this beutifal paper <https://pubmed.ncbi.nlm.nih.gov/26544944/>

Huge task; Note down what each Variable Means. If writing report this should go at the very top of the paper.

**Initial Questions:**

Now what?

PVCA related question: why does binned continuos variables capture more variation within the GEX data compared to raw variables?

Should I have completeley ommited the NAs from this data: IMO no this makes the data more realistic?

Would it be wise for Combat-Seq batch-correction? would this condense meta-variables to PC1 ? How would this change variance ratio of the PCS?

There is a paper that uses ConsenusClustering: <https://pubmed.ncbi.nlm.nih.gov/33903734/> (Likely next step to verify continuous latent variables) –\> Recc Reading: Han et al. 2021, and Wilkerson, Hayes (2021): Consensus Clustering implementation and how to interpret consensus matrices, track items and use CDF plots. for cluster stabiity

How should this data be labeled for the classifier ? If we do build one The focus will obviously be heavy on dim reduction utilizing the left unitary matrix

It's likely that we will overfit hard on this Data we'll likely need a mixture of probabilistic and LR + random Forest this. The decision boundary is going to be a pain the ass. I know somehow I'll end up using a deep-learning method.

I don't really have an end goal for this? Would it be to build a supervised classifier for prostate cancer data?

Demonstrate how anchoring a classifier to latent continuous variables allows ML to extract independent, clincally predictive GEX signatures

Possible Moves for the classifier: SVM with Radial Basis Function/ Gausian Kernel
